Opposing, spatially-determined epigenetic forces impose restrictions on stochastic olfactory receptor choice

Version 1

Abstract

Abstract

Olfactory receptor (OR) choice represents an example of genetically hardwired stochasticity, where every olfactory neuron expresses one out of ~2000 OR alleles in a probabilistic, yet stereotypic fashion. Here, we show that topographic restrictions in OR expression are established in neuronal progenitors by two opposing forces: polygenic transcription and genomic silencing, both of which are influenced by dorsoventral gradients of transcription factors NFIA, B, and X. Polygenic transcription defines spatially constrained OR repertoires, among which one OR allele may be selected for singular expression later in development. Heterochromatin assembly and genomic compartmentalization preferentially eliminate from this “privileged” repertoire ORs with more dorsal expression destinations, which are ectopically transcribed in neuronal progenitors throughout the olfactory epithelium. Our experiments identify early transcription as an “epigenetic” contributor to future developmental patterning and reveal how two spatially responsive probabilistic processes act in concert to establish deterministic, precise, and reproducible territories of stochastic gene expression.

Introduction

The development of multicellular organisms relies on gene expression programs that are precisely regulated in space and time. To transform probabilistic biochemical reactions, such as transcription and translation, into reproducible differentiation processes, plants and animals convert individual cellular variability into predictable cell population averages. Yet, there are cases in biology where gene expression variability is desirable, as it generates diverse cellular identities that are difficult to obtain with deterministic gene regulation. For example, production of antibodies via VDJ recombination, and evasion of immunological responses by antigenic variation represent biological systems that seek utmost randomness[1, 2]. Other biological functions, however, benefit from balancing absolute determinism with complete randomness, producing biased stochasticity. Genetically encoded biased stochasticity is often observed in the nervous system, where gene expression choices generated by neurons must integrate into functional and reproducible circuits[3]. In fly ommatidia, for example, biased randomness preserves a ratio of photoreceptor neuron identities across animals[4], whereas in mammals, random Protocadherin promoter choice[5], was recently shown to obey spatial patterns in the mouse neocortex, assuring proper tiling between neighboring neurons[6].

Mammalian olfactory receptor (OR) gene choice provides an extreme case of hardwired biased randomness[7]. OR transcription starts in neuronal progenitors of the main olfactory epithelium (MOE), which transiently express 5-15 ORs out of >1,000 OR genes distributed across chromosomes[8–10]. As these progenitor cells differentiate into post-mitotic olfactory sensory neurons (OSNs), they switch from polygenic to monogenic and monoallelic OR transcription[11]. This transition is mediated by the assembly of a multi-chromosomal enhancer hub over a single OR allele[12–14], followed by the stabilizing effects of an OR-elicited feedback signal[15–19]. During this developmental progression, heterochromatic silencing[20] and genomic OR compartmentalization[21, 22] act together to assure that the non-chosen OR alleles will remain transcriptionally repressed for the life of the OSN. Interestingly, the position of the OSN across the dorsoventral (DV) axis of the MOE predisposes this singular transcriptional choice towards a group of 50-200 OR genes[23], providing reproducible topographic restrictions in OR expression. The anatomical segments of the MOE that express a specific collection of OR identities are known as “zones”, with their total number varying from 4 to 9, depending on the analyses and criteria used to define them[24–27]. Although zonal restrictions in OR expression have a well-established influence in the wiring of the olfactory circuit[28, 29], the mechanisms that bias this singular transcriptional choice towards specific OR identities remain unknown.

Here, we identified genetically encoded mechanisms that introduce topographic biases in OR gene regulation. We report that OSN progenitor cells from various MOE segments transcribe OR mixtures consisting of ORs with the corresponding or with more dorsal expression identities. Ectopic expression of dorsal ORs at the polygenic state of OR transcription is rectified during differentiation by preferential genomic silencing that is skewed towards ORs with more dorsal expression identities than the identity of the OSN. Patterns of polygenic OR transcription and genomic OR silencing are influenced by gradients of transcription factors Nfi A, B, and X [30]. Triple Nfi deletion reduces heterochromatic silencing and genomic compartmentalization from dorsomedial ORs and eliminates them from ventral ORs, whose transcription in progenitor cells also becomes extinguished. Furthermore, spatial transcriptomics revealed a dramatic expansion of dorsomedial OR expression towards the ventral MOE and reciprocal transcriptional reduction of ventral ORs in triple Nfi cKOs, suggesting that patterns of genomic OR silencing and polygenic OR transcription influence OR gene choice. Indeed, transcriptional induction of an OR allele in OSN progenitors biases the choice of this allele in mature OSNs (mOSNs) throughout the MOE. Strikingly, by modulating the levels of OR induction in progenitor cells we can restrict expression of an OR allele in more dorsal OSNs, where heterochromatic silencing and genomic compartmentalization is less prevalent. Thus, our studies reveal that position-responsive OR transcription in OSN progenitors acts as an “epigenetic” signal for future singular choice among the previously transcribed ORs. Moreover, our data suggest that polygenic transcription and heterochromatic silencing/genomic compartmentalization act as opposing regulatory “rheostats” that determine in a spatially influenced fashion the exact OR repertoire that is available for stochastic singular choice in mature OSNs.

Results

OSN progenitors co-transcribe an increasing number of zonal OR identities towards the ventral MOE

The mouse MOE is divided into a limited series of stereotypic segments of OR expression that exhibit bilateral symmetry between the two nasal cavities (Figure 1A). In whole mount views, these segments present a dorsoventral (DV) segmentation pattern, with zone 1 being at the dorsal- and zone 5 at the ventral end of the MOE. Intricate invaginations of the MOE occurring during embryonic development and early postnatal growth, convolute this dorsoventral segmentation pattern, especially when viewing coronal sections of the MOE (Figure 1A). However, we will continue referring to the DV coordinates of each one of the five segments, or zones, as they correspond to their initial patterning during development.

Polygenic transcription of OR genes in olfactory progenitors follows a zonal expression pattern.

(A) Schematic illustrating OR zones along the dorsoventral axis, in whole mount views of the MOE (left) and coronal sections (middle). Zone1 (red) is the dorsal-most zone and zone5 (blue) is the ventral-most zone. Zoomed in view of a coronal sections (right) depicts MOE cell populations from different stages of OSN differentiation organized in a pseudostratified fashion from the basal (least differentiated) to apical (most differentiated) layers: HBC, horizontal basal cell; GBC, globose basal cell; INP, immediate neuronal precursor; iOSN, immature olfactory sensory neurons; mOSN, mature olfactory sensory neuron.

(B) t-SNE clustering of single cells from FAC-sorted cell populations with Seurat based on the most variable genes showing the separation of single cells into 6 populations (left panel). The 6 populations were assigned cell identities using expression of known MOE markers[31] (See also Figure S1). Olfactory receptor expression is first detected in INP3 cells (right panel).

(C, D, left panel) t-SNE clustering from FAC-sorted cell populations isolated from dorsal (zones 1-2 in C) or ventral (zones 3-5 in D) MOE microdissections. Cells are colored according to the zonal index of the most highly expressed OR. (C, D, right panel) Plots depicting zonal identities of all the ORs detected in individual INP3 cells from dorsal (C) or ventral (D) microdissections. Y axis shows OR expression in normalized UMI counts for different ORs (separated by black lines). On the X axis each point is a different INP3 cell. ORs are colored according to their zonal index.

(E) OR expression by zonal index in olfactory progenitor INP cells (top) and mOSNs (bottom), determined with bulk RNA-seq, in cells isolated from dorsal-most (zone 1) (left), dorsomedial (zones 2-3) (middle), and ventral-most (zones 4-5) MOE microdissections (right). Note that INP and mOSN cells were FAC-Sorted from the same exact dissection, thus the mOSN OR expression patterns confirm the accuracy of the dissection.

Within each zone mOSNs express a single OR allele that is randomly chosen among 50-200 OR genes with proper zonal identities. However, before the onset of singular OR expression, mitotically active OSN progenitors, the immediate neuronal precursor (INP) cells, co-express, multiple lowly expressed ORs [8–10]. To determine whether zonal restrictions are operational from this polygenic stage of OR transcription, we performed deep scRNA-seq analyses of FAC-sorted OSNs and OSN progenitors from micro-dissected dorsal or ventral MOE segments (Supplemental Figure S1A). To enrich our plate-based scRNA-seq for cell populations of interest we used Mash1CreER; tdTomato; Ngn1-GFP triple transgenic mice (Supplemental Figure S1B). We injected P4 mice with tamoxifen, inducing permanent tdTomato expression, and then collected cells 48 hours later (Supplemental Figure S1B). From each dissection we isolated four major cellular populations corresponding to four successive differentiation stages, as previously described[31, 32]: GBCs (MOE stem cells), INPs (immediate neuronal precursors), iOSNs (immature OSNs), and mOSNs (Supplemental Figure S1C-D).

We first detect OR mRNAs in INP3 cells (Figure 1B), which consistently transcribe multiple ORs. Surprisingly, while dorsal INPs transcribe almost exclusively dorsal ORs, ventral INPs have complex OR mixtures consisted of every zonal identity (Figure 1C-D). With a 3 UMI cut off, we detect dorsal ORs in 43 ventral INP cells, and ventral ORs in only 29 of them, while dorsal INP cells express predominantly dorsal ORs (Figure 1C-D). Moreover, as the ventral INPs differentiate to iOSNs, dorsal OR transcription is replaced by ventral, zone-appropriate OR transcription, culminating to singular expression of an OR allele with the correct zonal identity (Figure 1D, Supplemental Figure S1). These observations were independently confirmed by bulk RNA-seq on FAC-sorted INP and mOSN cells extracted from trisected dorsal, medial, and ventral MOE. This bulk analysis showed that in every case INPs co-transcribe ORs with the correct and with more dorsal zonal identities, while further differentiation replaces dorsal ORs with ORs of the correct identity (Figure 1E). This immediately poses mechanistic questions about the process that shuts off dorsal ORs and enhances the transcription of the ORs expected to be expressed in each MOE segment.

Heterochromatin eliminates ectopically expressed ORs along the dorsoventral MOE axis

We previously showed that OSN differentiation coincides with heterochromatin-mediated OR gene silencing [20]. If heterochromatinization contributes to singular OR choice by eliminating every non-chosen OR transcribed in INPs, then in any MOE segment silencing should be preferentially applied to ORs expressed in that segment and to more dorsal ORs. Therefore, dorsal ORs (zone 1 ORs), which are expressed in INPs throughout the MOE, should have highest levels of heterochromatin, whereas ventral ORs (zone 5 ORs), transcribed only in ventral INPs, should have the lowest, with the rest of the OR repertoire having intermediate levels of heterochromatic marks. Visual inspection of ChIP-seq genomic tracks along OR gene clusters with mixed zonal constitution, reveals highest H3K9me3/H3K79me3 levels on the dorsal ORs and lowest on the ventral ORs of the cluster (Figure 2A). Aggregate ChIP-seq analysis of all the ORs grouped by zonal identities, corroborates the gradual reduction of H3K9me3 and H3K79me3 enrichment from dorsal to ventral ORs for the whole OR repertoire (Figure 2B). The only exception from this pattern is found at the dorsally expressed class I ORs, which rely on different regulatory mechanisms than the canonical class II ORs [33, 34] (Figure 2B). Finally, using the FACS-based strategy described earlier, we confirmed that both heterochromatic marks are predominantly deposited during the INP to iOSN transition, simultaneously with the transition from polygenic to singular, zonally appropriate OR expression (Figure 2C, Supplemental Figure 2A). Importantly, the descending pattern of heterochromatin enrichment from dorsal to ventral OR identities is preserved throughout differentiation.

Heterochromatin deposition silences OR genes from lower zones.

(A) Signal tracks of H3K9me3 and H3K79me3 native ChIP-seq from the whole MOE over two representative OR gene clusters that were selected because they harbor ORs from both zone 1 and zone 5. Below the signal track OR genes are colored according to their zonal index: zone1 ORs in red, zone2 ORs in yellow, zone 5 ORs in blue and ORs with unknown zonal index in gray.

(B) H3K9me3 (left) and H3K79me3 (right) native ChIP-seq in the whole MOE. Box plots of read density over OR gene bodies separated by their zonal index depict a pattern of deposition that is high on dorsal-most expressed zone 1 OR genes, progressively decreases with more ventral zonal OR indexes, and is absent on class I ORs.

(C) H3K79me3 native ChIP seq in GBC, INP, iOSN and mOSN populations shows onset of H3K79me3 deposition in INP cells. Each row of the heatmaps shows coverage over an OR gene body (separated into categories by their zonal index). (See also Supplementary Figure S2 for H3K9me3 heatmap).

(D) H3K79me3 native ChIP-seq in mOSNs from zonally dissected MOE. Colored schematics above each heatmap depicts the zone of dissection. (See also Supplementary Figure S2 for H3K9me3 heatmap).

We then asked if the patterns of heterochromatin deposition detected in mixed OSNs from the whole MOE are preserved in distinct zones. ChIP-seq in dorsal, dorsomedial, and ventral mOSNs showed that OR groups transcribed in INPs have higher heterochromatin levels than the ones that were not transcribed in that MOE segment (Figure 2D, Supplemental Figure S2B). Thus, most ORs are heterochromatic in ventral OSNs; dorsal and dorsomedial ORs are heterochromatic in dorsomedial OSNs; and only dorsal ORs have some heterochromatin in dorsal OSNs (Figure 2D). Although each zonal OR group is heterochromatic in the MOE segment that is expressed, it has lower enrichment of H3K9me3/H3K79me3 in its own segment than in more ventral segments, where it not chosen for stable expression. Thus, dorsal ORs have less heterochromatin in dorsal OSNs than the rest of the MOE, and dorsomedial ORs have less heterochromatin in dorsomedial OSNs than ventral OSNs. Similarly, at the ventral end of the DV axis, ventral ORs have less heterochromatin than dorsal and dorsomedial ORs. Detection of heterochromatin on ORs from the zonal group that is expressed is not counterintuitive, as only one OR allele from the ones co-transcribed will be eventually chosen, and the rest must be silenced (Supplemental Figure S2C). In other words, in every MOE segment, OR heterochromatinization is preserved only for the ORs that have the potential to be expressed and is not applied to more ventral ORs, which were not transcriptionally active in INPs. This is consistent with recent reports of heterochromatin marks being detected on trace amine-associated receptor (TAAR) genes only in TAAR-expressing OSNs and not the rest of the MOE (REF).

DV gradient of OR gene compartmentalization follows patterns of heterochromatin assembly

Heterochromatic ORs converge into multi-chromosomal genomic clusters of extreme chromatin compaction that contributes to the effective and stable OR silencing [21]. We thus asked if the spatially determined pattern of OR heterochromatinization at the linear genome coincides with similar patterns of 3D genomic compartmentalization. In situ Hi-C in FAC-sorted OSNs from MOE segments along the DV axis revealed a striking resemblance between deposition of heterochromatic marks and genomic compartmentalization. For example, inspection of the genomic interactions between 3 OR clusters in chromosomes 2, shows that a cluster of ventral ORs is recruited to OR compartments only in ventral OSNs, where they are heterochromatic (Figure 3A). In contrast, the other two clusters, which are either enriched for dorsal ORs, or have mixed constitution, make strong genomic contacts with each other in all three MOE segments (Figure 3A). To expand this analysis to every OR, we measured the frequency of interchromosomal genomic interactions between ORs with different DV identities, reaching the same conclusion: Interactions between dorsal ORs is the default in every OSN, whereas compartmentalization for the remaining of the OR repertoire increases along the DV axis (Figure 3B). Intriguingly, as with levels of heterochromatin, we detect the following recurrent pattern of OR compartmentalization: every OR has intermediate HiC contact frequencies with other ORs in their expression segment, lower HiC contact frequencies in more dorsal segments, and higher HiC contacts ventrally.

Zonal OR compartmentalization permits OR genes from more ventral zones to be recruited into the OR compartment.

(A) In situ Hi-C contact matrices of a 90Mb region of chromosome 2 that contains 3 large OR gene clusters, depicted with orthogonal boxes under the contact matrices. HiC libraries were prepared from mOSNs FAC-sorted from dorsal-most (zone1), dorsomedial (zone2) and ventral-most (zone 4/5) MOE microdissections. For each contact matrix, we magnify the long-range cis contacts between the large OR cluster that contains ORs from every zone with the OR cluster that contains mostly zone 1-2 ORs (red box on the left) and the OR cluster that contains mostly zone 4-5 ORs (blue box on the right). All cis OR contacts increase from dorsal to ventral OSNs, but zone 4/5 ORs associate with the other ORs only in the most ventral OSNs (as noted by the HiC signal on the blue box).

(B) Heatmaps of average interchromosomal Hi-C contacts between OR genes annotated by their zonal index at 50Kb resolution shows increased trans contacts in OSNs from more ventral zones. OR genes have a similar, intermediate frequency of contacts in the mOSN population where they are expressed, marked with an asterisk.

(C) Dip-C on mOSNs from dorsal and ventral dissected MOE was used to generate haplotype resolved single cell contact matrices and 3D genome structures, as previously described[37].

(D) Analysis of Dip-C contact densities of interchromosomal contacts between ORs genes confirms that ventral mOSNs have increased OR compartment interactions (Wilcoxon rank sum p-value = 9.164e-11).

(E) Single cell heatmaps of pairwise distances between OR genes generated from 3D genome structures in two ventral mOSNs show OR genes from different chromosomes intermingle in a different pattern in the two cells (top). For each cell, heatmaps are sorted by chromosome order and show all OR interactions within 10 particle radii (approximately ~600nm). Representative 3D structures showing the different positioning of three chromosomes (chr19, chr9 and chr2) in the two cells resulting in a different pattern of OR cluster contacts (bottom). See also supplemental Figure S3 for a complete map of all the Dip-C libraries.

The “intermediate” levels of heterochromatin enrichment and HiC contacts observed on ORs within their expression zone may reflect a less compact, transcription-compatible state of heterochromatin, or less frequent silencing of these ORs compared to more dorsal ORs. To distinguish between the two scenarios, we explored OR silencing at the single cell level using Dip-C[35–37]. We performed Dip-C in 48 dorsal and 48 ventral OSNs (Figure 3D). We used haplotype resolved data to compute distances of all genomic loci at 20kb resolution and generated 3D models for all cells (Figure 3C), as previously described[37]. Analyzing contact densities between OR loci, as well as distances in the 3D model we confirmed that OR compartments from ventral OSNs are larger and contain more ORs from more chromosomes than in dorsal OSNs (Figure 3D), consistent with our bulk HiC data. Importantly, “stronger” HiC contacts among dorsal ORs observed in bulk, represents increased number of dorsal ORs participating in OR compartments in each OSN, rather a closer distance between dorsal ORs within a compartment (data not shown). Thus, extrapolating Dip-C results to H3K9me3/H3K79me3 enrichment, we conclude that “intermediate” silencing levels of each OR group in their own zone likely reflects less frequent silencing than ORs from a more dorsal zone. In this note, OR compartmentalization is highly probabilistic, with each one of the 48 dorsal and ventral OSNs having unique maps of OR-OR genomic interactions (Figure 3E) (Supplemental Figure 3A, B). Thus, we propose that the balance between two probabilistic, yet DV-responsive processes, early transcription and genomic silencing defines the OR ensemble that is available for singular choice along the DV axis. To test this model, we sought to identify factors responsible for generating these remarkable patterns.

NFI paralogues generate DV patterns in OR expression

We searched for transcription factors that have strong expression during the INP to iOSN transition that is graded across the DV axis of the MOE. NFI paralogues NFIA, B, and X have strong, DV-influenced expression in INPs that is preserved in iOSNs (Figure 4A, B and Supplemental Table 1). Specifically, NFIA and NFIB are expressed higher in ventral INPs and iOSNs, and NFIX is higher in ventral mOSNs (Figure 4B).These three members of the nuclear factor I (NFI) family of transcription factors control a plethora of developmental and cell specification processes[30, 38], and were previously implicated in OSN differentiation[39, 40]. Thus, we decided to explore genetically their contribution in the establishment of dorsoventral patterns of OR expression.

NFI paralogue gradients regulate zonal OR expression

(A) Heatmaps showing the expression levels of differentially expressed transcription factors between dorsal, dorsomedial, and ventral cells during the INP and iOSN developmental stages. The shown transcription factors are significantly differentially expressed between dorsal and ventral cells with an adjusted p-value of <0.05, at least a three-fold change in expression, and an expression of at least 15 TPM. A broader list of transcription zonal factors is included in Table S1. The heatmaps is sorted based on expression in ventral cells and a color bar above each heatmap shows the log2 fold change in ventral cells relative to dorsal cells.

(B) Expression levels of NFIA, NFIB, and NFIX at four stages of OSN development in dorsal cells (red), dorsomedial cells (green) and ventral cells (blue).

(C, D) Comparison of OR gene expression in NFI ABX triple knockout and control cells from the whole MOE. NFI transcription factors are deleted either in progenitors (C) using a KRT5-CreER driver or in mOSNs (D) using an Omp-IRES-Cre driver (as described in Figure S4). At the schematic depiction of the experimental strategy at the right of each panel, scissors indicate the differentiation stage of the deletion, and red box the cell type that was FAC-sorted for RNA-seq analysis.

(E) OR expression in NFI ABX triple knockout, Nfi AB double knockout, Nfi X knockout and control mOSNs from ventrally dissected MOE. Knockout was induced in progenitors with KRT5-CreER. Plots show a different pattern of OR gene transcription in the different genotypes. Quantification of differentially expressed ORs for each genotype shown in Figure S4.

To interrogate the potential role of NFI A, B and X in zonal OR expression we deleted all three genes simultaneously using the Krt5CreER driver, which is expressed in quiescent stem cells of the MOE (HBCs). We crossed Krt5CreER; tdTomato mice to NFIA, B, X fl/fl mice[41], and induced recombination with tamoxifen. To force the quiescent HBCs to differentiate into OSNs, we ablated the MOE with methimazole and allowed 40 days for a complete restoration by the marked progeny of the NFI KO or control HBCs (Supplemental Figure S4A), as previously described[14]. RNA-seq analysis of the FAC-sorted OSNs from the whole MOE, revealed significant transcriptional reduction of ventral OR identities and reciprocal increase of dorsomedial ORs, with the transcription of the dorsal-most ORs being unaffected (Figure 4C). In contrast, triple NFI deletion only in mOSNs, with OMPiresCre has no measurable effects in OR expression (Figure 4D). To ask whether the reduced transcription of ventral ORs reflects a developmental defect of ventral OSN differentiation, versus a bona fide dorsalization of ventral OSNs, we performed RNA-seq in OSNs isolated specifically from ventral MOE microdissections. This experiment revealed ectopic expression of dorsomedial OR identities in place of the proper ventral ORs (Figure 4E), a result confirmed by immunofluorescence (IF) experiments (Supplemental Figure S4B,C). This transcriptional transformation of ventral OSNs satisfies the original criteria of homeosis [42], since the overall mOSN identity is not altered by the triple NFI deletion: Only 13 out of ~200 OSN-specific genes are significantly different between control and KO OSNs, and 117/207 non OR zone 4/5 enriched genes are still expressed in the ventral-most zones, acting as independent fiducial markers for our zonal dissection (Supplemental Figure S4E). The severity of this dorsomedial transformation depends on the number of NFI genes deleted, with the triple cKO expressing predominantly zone 2 and 3 ORs, double NFIA, B cKO expressing zone 3 and 4 ORs, and single NFIX cKO having almost wild type expression patterns of zone 4 and 5 ORs (Figure 4E, Supplemental Figure S4D).

Spatial transcriptomics reveal widespread homogenization and dorsalization of the MOE upon triple Nfi deletion

To obtain a complete and unbiased understanding of the consequences of triple Nfi deletion in patterns of OR expression we deployed a spatial transcriptomic approach. Since our goal was to decipher zonal patterns of OR expression across the dorsoventral MOE axis without requirements for single cell resolution, we opted for the Visium Spatial Gene Expression workflow (10X Genomics)[43]. This workflow is ideal for interrogation of spatial OR expression in mOSNs, as OR mRNAs are highly abundant and readily detectable in most spatial spots that contain OSN mRNAs. For increased stringency, we only included spatial spots that include more than 2 OR genes and 3 OR transcripts. We analyzed 4 MOE sections from NfiA,B,X triple cKO mice and littermate controls, from two mice each (Figure 5A). Expression data on OR genes were normalized and integrated across replicates (see STAR methods). We performed PCA analysis, by which spatial spots were arranged in 5 clusters in control and cKO MOEs (Figure 5B). Interestingly, while dimensionality reduction and unbiased clustering generated OR clusters that correspond to zonal patterns of OR expression, i.e., each cluster contains ORs from a specific zone in control MOEs, only zone 1/class I ORs followed this correlation in cKO MOEs (Figure 5B). The other 4 clusters homogenously express zone 2-4 ORs, with expanded expression of zone 2 ORs in every cluster and loss of zone 5 OR detection. Thus, conditional triple Nfi deletion causes loss of spatial patterning for zone 2-4 OR genes and loss of expression for zone 5 ORs, without influencing the expression of zone 1 ORs.

Spatial transcriptomics shows dorsalization and homogenization of the MOE upon Nfi A, B, and X deletion.

(A) Schematic depicting our analysis pipeline: Spatial transcriptomics was performed on sections of control (wt) and Nfi ABX cKO MOE. Dimensionality reduction was performed, and spatial spots were clustered based on normalized expression of OR genes.

(B) Heatmaps showing scaled, normalized expression levels of the top 20 highest expressed OR genes per zone in the control (wt) dataset. Unbiased neighborhood analysis and clustering grouped spatial spots into 5 clusters for both control (wt) and cKO MOE (depicted in distinct colors on the top of the heatmaps). Clustering of spatial spots in control (wt) samples reproduces anatomical zones, as spots within each cluster express OR genes with the corresponding zone index (left heatmap). We generated the same clusters in Nfi cKO samples (right heatmap). Although cluster 1 expresses exclusively zone 1 ORs, like in control MOEs, clusters 2-5 exhibit homogenous OR expression, with ventral expansion of zone 2/3 ORs, and reduced representation of zone 4/5 ORs.

(C) Average normalized per-spot expression of the 20 highest expressed OR genes from zone1, zone2, and zone5 is overlaid against H&E histological image of control (wt, top) and Nfi cKO (bottom) MOE sections. Expression of zone1 OR genes is confined to the same anatomical region for both control and Nfi cKO sections. Zone2 OR gene expression is spread to more ventral regions in the Nfi cKO compared to control sections. Expression of zone5 OR genes is detected in fewer spatial spots and at a lower expression level in the knockout sample.

(D) Spatial spots were assigned to a zone based on the highest summed normalized expression of OR genes. Zonal spot assignment of the control (wt) sample visually reproduces known anatomical zones. In the Nfi cKO sample, spots in the dorsal region had the highest expression of Class I and zone1 OR genes, similar to control sample. However, at the rest of the Nfi cKO MOE, most spots have a zone 2 OR identity. Spots assigned ‘None’ did not contain any OR transcripts and were excluded from cluster analysis.

To depict the effects of triple Nfi deletion in spatial patterns of OR expression we plotted the average OR expression per spatial spot of the top 20 most highly expressed OR genes for zones 1, 2, and 5. We then overlaid the corresponding values against the histological images of the control (wt) and Nfi ABX cKO MOEs (Figure 5C). As observed in the clustering and heatmap analysis, zone1 OR expression is confined to the same anatomical region for both samples. However, zone 2 OR expression in the cKO MOE extends beyond its defined anatomical region of the control MOE, and spreads to the ventral-most zones (Figure 5C). This expansion is also observed in the expression of individual zone 2 genes (Supplemental Figure S5A). In contrast, the top 20 zone 5 OR genes, while highly expressed in control MOEs, are almost undetectable in Nfi cKO MOEs (Figure 5C), consistent with our RNA-seq analysis. Olfr1507, the most highly expressed zone 5 OR, is undetectable in the cKO spatial spots (Supplemental Figure S5B), in agreement with our IF data. Finally, to obtain a more general understanding of the spatial transformations in OR expression patterning upon triple Nfi deletion, we assigned a zonal index of each spatial spot using the maximum normalized expression of all the OR genes detected in a spot (see STAR methods). Unlike control MOEs, where spot assignment reproduces zonal anatomical positions, most spatial spots in the cKO MOEs are assigned to zone 2, in a striking dorsalization and homogenization of the MOE (Figure 5D).

NFI gradients control patterns of OR heterochromatinization and polygenic OR transcription

We searched for a mechanistic explanation for the homeotic transformation of ventral OSNs in NFI cKO mice. Our experiments so far have identified 3 spatially responsive processes that may contribute to the dorsoventral patterning of OR gene choice: polygenic OR transcription in INPs, OR heterochromatinization and genomic compartmentalization during INP to iOSN transition. Thus, we explored the effects of triple NFI deletion in all three processes. First, we investigated the effects of NFI deletion in OR heterochromatinization by ChIP-seq and HiC on triple NFI cKO OSNs from the ventral-most MOE segments. ChIP-seq revealed an almost complete loss of heterochromatin from ventral ORs and small reduction in dorsomedial ORs in NFI cKO ventral OSNs (Figure 6A, Supplemental Figure S7E, F). Similarly, in situ Hi-C in control and triple Nfi KO OSNs from ventral MOE segments, revealed a strong reduction in the long-range cis and trans genomic contacts made by ventral ORs, and intermediate reduction for dorsomedial ORs (Figure 6B). Dorsal ORs, which are not affected transcriptionally, did not exhibit any change in ChIP-seq and HIC contacts. (Figure 6A, B). Strikingly, in both processes, heterochromatin assembly and genomic compartmentalization, the patterns observed in ventral OSNs upon Nfi deletion, are like the ones observed in dorsomedial OSNs from the control MOEs (Supplemental Figure S6A, B).

Nfi A, B and X regulate chromatin state and OR compartment formation

(A) Native ChIP-seq for H3K9me3 (top) and H3K79me3 (bottom) in Nfi ABX knockout mOSNs from ventral MOE. Heatmaps of ChIP signal over OR genes show decrease of both histone marks on zone 3-5 index ORs in NfiABX knockout compared to control. Triple Nfi deletion was induced by KRT5-CreER (before OSN differentiation).

(B) HiC in in Nfi ABX knockout mOSNs from ventral MOE. Left: In situ Hi-C contact matrices of a 90Mb region of chromosome 2 from control (top) and NFI ABX triple knockout (bottom) mOSNs, as described in Figure 3A. from ventral MOE show long-range cis interactions between 3 large OR gene clusters. Note that long range cis contacts the zone 4/5 enriched cluster and the mixed cluster dissipate in the triple Nfi cKO (bottom blue box), whereas the contacts of the mixed cluster with the zone 1/2 enriched cluster are preserved (bottom red box). Right: Heatmaps of average interchromosomal Hi-C contacts between OR genes annotated by their zonal index (as described in Figure 3B) in control (top) and triple Nfi cKO (bottom) mOSNs from ventral MOE. Trans contacts between zone 4/5 ORs dissipate, whereas trans contacts between zone 2/3ORs reach intermediate levels detected at “in zone” ORs.

(C) OR expression by zonal index in INP cells isolated from ventral Nfi ABX knockout and control MOE. Nfi ABX INP cells were isolated as described in Figure S4. Log2FC of OR expression in Nfi ABX vs control INP cells, shows a decrease in expression of zone4-5 ORs (right).

Finally, we explored the effects of triple Nfi deletion to the polygenic transcription of ORs in INP cells. We used a FACS-based strategy to isolate INPs from the ventral MOE followed by bulk RNA-seq as described earlier. Again, as with the results from ChIP-seq and HiC experiments, we detect a conversion toward the signatures observed in dorsomedial INPs, i.e., detection of only dorsal and medial ORs and depletion of ventral OR identities from the INP transcriptome (Figure 6C). Thus, our data reveal an unexpected correlation between OR transcription in INP cells, and two diametrically opposing gene expression outcomes in OSNs: silencing for the majority of the co-transcribed OR alleles and singular choice for one of them. Although transcription mediated silencing is widely deployed for the stable repression of repetitive elements, transposons and retroviruses that integrate in the genomes of various organisms, a transcription-mediated gene choice mechanism that transcends differentiation stages is uncommon. Thus, we devised a genetic strategy that would test the hypothesis that polygenic OR transcription is a pre-requisite for singular OR choice.

Early OR transcription promotes OR gene choice in mOSNs

We manipulated OR transcription using a genetically modified Olfr17 allele with a tetO promoter inserted immediately downstream of its transcription start site[44]. This allele enables strong transcriptional activation of Olfr17 from the endogenous genomic locus under the control of tTA (Figure 7A, Supplemental Figure S7A). Olfr17 expression is monitored by an iresGFP reporter inserted immediately downstream of the Olfr17 translational stop codon (Figure 7A). To induce transcription of this tetO-Olfr17iGFP OR allele in INPs and iOSNs, we used Gng8-tTA transgenic mice. Gng8 is expressed predominantly in INPs, downregulated in iOSNs, and completely shut off in mOSNs. Consistent with the expression properties of Gng8 and previous reports[45, 46], we only detect GFP in the basal MOE layers of Gng8-tTA; tetO-GFP mice (Figure 7B), which are enriched for INPs and iOSNs. However, when we cross the same Gng8-tTA driver to tetO-Olfr17iGFP mice, we detect widespread GFP signal in apical MOE layers, which contain predominantly mOSNs (Figure 7B). Since there is no tTA expression in mOSNs, we reasoned that the INP/iOSN-induced tetO-Olfr17iGFP allele is chosen for expression by the endogenous transcriptional machinery responsible for singular OR choice. Indeed, HiC experiments of these OSNs revealed that Greek Islands, the intergenic OR enhancers that converge over the chosen OR allele[12–14], are recruited specifically to the tetO-Olfr17iresGFP allele (Figure 7C), explaining the sustained expression of this OR in mOSNs. Furthermore, high doxycycline diet to these mice for 30 days fails to extinguish tetO-Olfr17iresGFP expression in mOSNs (Supplemental Figure 7), in support of the notion that transcriptional induction of Olfr17 in INPs/iOSNs, signals for the preferential choice of this OR in mOSNs. In contrast, ChIP-seq experiments on GFP-negative OSNs, which do not express this OR allele, did not reveal a significant increase of heterochromatin on the P2 locus (data not shown).

Genetic induction of OR transcription in olfactory progenitors determines OR choice in mOSNs

(A) Genetic strategy for transcriptional induction of OR Olfr17 (a zone2 index OR) from its endogenous genomic locus. A genetically modified “tetoOlfr17” allele contains a tetO promoter immediately downstream of the endogenous Olfr17 promoter and an IRES GFP reporter after the coding sequence[44]. In the presence of tTA a high level of tetoOlfr17 is induced from the tetO promoter (top), while in the presence of a high amount of doxycycline tTA is inhibited and transcription is regulated by the endogenous promoter. See also Supplemental Figure S7A for information of the genomic locus of this Olfr17 allele.

(B) tTA driven by the Gng8 promoter is expressed in INP and iOSN cells in the MOE[52]. When Gng8-tTA drives expression of tetoGFP transcription is driven only in progenitor cells located on the basal side of the MOE, where the tTA is expressed (left)[46]. In contrast, when Gng8-tTA drives expression of tetoOlfr17, expression persists in mature OSNs where tTA is no longer present (right). See also Supplemental Figure S7A,B for the sustained and widespread expression of this OR allele after 30 days in DOX treatment and for scRNA-seq data on Gng8 expression during OSN differentiation.

(C) In situ Hi-C in tetoOlfr17 expressing cells shows enriched contacts with interchromosomal Greek Island enhancers over the Olfr17 locus, suggesting tetoOlfr17+ mOSNs are using endogenous mechanisms to sustain Olfr17 expression.

(D) tetoOlfr17 expression in coronal sections of the MOE determined by GFP fluorescence. In the absence of tTA induction tetoOlfr17 expression occurs only in zone2 of the MOE (right); with high tTA induction in progenitor cells expression occurs throughout all zones of the MOE (left); and with low tTA induction in progenitors, due to the addition a low amount of doxycycline, expression occurs in zone2 and spreads dorsally to zone1 (middle). Zoom in shows tetoOlfr17 expression in its native zone2 (i) and ectopic expression in the most ventral zone 5 (ii). Mice on low doxycycline treatment were provided doxycycline at 1ug/ml in water throughout gestation and postnatal life. See Supplemental Figure S7C for the Dox administration protocol.

(E) Quantification of tetoOlfr17 expression relative to a normalized zonal position (illustrated on the left) in tetoOlfr17 (bottom), tetoOlfr17 with Gng8tTA driver (top), and tetoOlfr17 with Gng8-tTA driver on low doxycycline (middle). 6 sections from two replicates were analyzed for tetoOlfr17 with Gng8-tTA; 9 sections from two replicates were analyzed from tetoOlfr17 with Gng8-tTA and low doxycycline; 29 sections from two replicates were analyzed for tetoOlfr17 without tTA. The plot displays a maximum of 1000 cells randomly selected for each condition.

(F) H3K9me3 native ChIP signal over the Olfr17 locus in mOSNs from dorsal MOE (red), dorsomedial MOE (green) and ventral MOE (blue) shows a higher level of heterochromatin in ventral MOE.

(G) Model of OR choice in each zone of the MOE, regulated by the interplay of low level polygenic transcription in INP cells which defines the OR repertoire that can be chosen in each zone, and heterochromatic silencing, which prevents ectopic expression of more dorsal ORs. See also Supplemental Figure S7D, E for ChIP-seq analyses on zone 4/5 INP cells from control and Nfi cKO INP cells.

Intriguingly, transient induction of Olfr17 transcription promotes preferential choice of this OR throughout the MOE, rather than only in zone 2, were Olfr17 is normally expressed (Figure 7D). In fact, the vast majority of mOSNs from zones 1 to 4 are GFP+, and only in the ventral-most zone 5 we detect a more sporadic pattern of ectopic Olfr17 choice (Figure 7D, E). We hypothesized that reduced frequency of ectopic Olfr17 expression in the most ventral segment reflects the fact that heterochromatin levels and genomic compartmentalization of this dorsomedial OR allele is highest at this MOE segment even at the INP stage (Supplemental Figure S7E). This immediately suggests that the balance between transcriptional activation heterochromatic silencing during INP to iOSN transition determines whether an OR can be chosen for singular expression. If this hypothesis is correct, then reducing Olfr17 transcription in INP/iOSN cells, should preferentially prohibit ectopic Olfr17 expression in ventral MOE segments, where heterochromatic silencing is stronger. To test this, we pharmacologically manipulated tTA activity using low levels of doxycycline (1mg/Kg) administration throughout the life of the mouse (Supplemental Figure 7A), which reduce but do not eliminate tTA-driven transcription. Remarkably, mice that were subjected to this doxycycline regimen, continue to frequently express Olfr17 in dorsal mOSNs (zones 1-2), but not in mOSNs from more ventral MOE segments (zones 3-5) (Figure 7D, E), where heterochromatin levels on this OR allele are highest (Figure 7F). Thus, we can manipulate the zonal expression of an OR allele in mOSNs, by pharmacologically modulating the frequency and levels of transcriptional activation in INP/iOSN cells.

Discussion

We uncovered a mechanism by which a random transcriptional process becomes skewed towards specific outcomes, transforming the relative position of a neuron across the dorsoventral axis of the MOE into biased OR gene choice. The solution to the perplexing segmentation of the MOE into distinct and reproducible territories of OR expression is the following: polygenic OR transcription in neuronal progenitors highlights a small group of ORs that can be chosen for singular expression later in development (Figure 7G). In each MOE segment this OR mixture includes ORs that should be expressed in mOSNs of the segment, as well as ORs that are only expressed in more dorsal MOE segments (Figure 7G). As these progenitor cells differentiate into iOSNs, heterochromatic silencing preferentially decommissions from this mixture more dorsal ORs, and with lower efficiency ORs that could be expressed in the segment, biasing this singular choice towards a spatially appropriate OR repertoire (Figure 7G). Our scRNA-seq analysis revealed two vectors in the determination of the OR ensemble that is co-transcribed in each OSN progenitor: chance, as every OR combination is unique, and determinism, as the overall zonal identities of the co-transcribed OR mixtures are informed by the position of the progenitor cell. Similarly, Dip-C revealed that genomic silencing also follows skewed patterns, eliminating preferentially ORs with more dorsal expression signatures than ORs that could be expressed in each zone. The final product of these opposing probabilistic “rheostats” is the generation of gene expression programs that may not have sufficient resolution to determine which OR will be chosen in every OSN but are precise enough to generate reproducible expression territories for each one of the ~1400 OR genes.

We identified gradients of transcription factors Nfi A, B, and X as partial orchestrators of the dorsoventral patterning of OR expression, which they establish as follows: they contribute to the silencing of dorsomedial ORs (zone 2 and 3 ORs); they activate both polygenic transcription and silencing of ventral ORs (zone 4 and 5 ORs); and they have no influence on the expression of dorsal-most ORs (class I and Zone 1 ORs). Given that Nfi factors are predominantly known as regulators of embryonic and adult stem cell biology[41, 47], it is surprising that in the olfactory system their deletion does not interfere with the maintenance of stem cell populations, but with the OR expression patterns in post-mitotic, fully differentiated mOSNs. Interestingly, triple NFI deletion after the onset of singular OR choice, has no effect in OR patterning, consistent with the emerging model that OR specification takes place exclusively at the INP to iOSN transition, and the notion that these patterning factors are not required for maintenance of OR transcription. Thus, we speculate that singular OR gene choice in OSNs can be executed by the common nucleoprotein complex of Lhx2/Ebf/Ldb1 bound to the multi-enhancer hub, consistent with the fact that we detect hubs of similar constitution associating with active ORs in different zones[14].

A question emerging from these observations is why not use the same transcription factor gradients to regulate both polygenic and monogenic OR transcription? The answer is likely related to the absolute requirement for transcriptional singularity: transcription factor gradients can transcribe specific OR mixtures in a DV-responsive fashion, but they cannot activate only a single OR promoter among the many they can bind to. But even if singularity was achievable by transcription factor combinations and the OR-elicited feedback, OR promoters with the strongest binding motifs would be consistently chosen first, excluding ORs with weaker promoters in a “winner takes all” model. This would result in preferential choice of specific ORs, reduced diversity in OR representation, and a narrower sensory spectrum for the olfactory system. With the process revealed here, the most frequently activated OR promoter in INPs/iOSNs has only a 5-10% chance of being chosen for singular expression in OSNs. Thus, by segregating OR gene regulation into two stages, polygenic transcription in progenitor cells and singular choice in OSNs, the olfactory system can impose deterministic biases while assuring equitable receptor representation. Of course, this system has limitations in preserving transcriptional equity: artificial transcriptional induction of an OR allele in OSN progenitors under the powerful tetO promoter, bypasses these constraints and results in biased choice of this allele in most mOSNs. This immediately suggests that cis OR regulatory elements are subject to selective pressure that preserves their weak transcriptional activation properties, explaining why robust OR transcription in mOSNs requires assembly of interchromosomal multi-enhancer hubs.

In this note, zones may also have evolved to satisfy the requirement for distributed OR representation: if dorsal-most ORs, which are detected in every OSN progenitor regardless of DV origin, have the most frequently activated promoters, then silencing them in more ventral MOE segments assures that other OR identities will also have the chance to be expressed. Consistent with this model, is the observation that mutations on the Lhx2 or Ebf binding sites of the promoter of dorsal OR M71, result in less frequent and more ventral M71 expression patterns[48]. Thus, DV segmentation of the MOE may serve as a mechanism that prevents ORs with stronger differences in promoter strength from competing for singular expression, assuring that every OR is expressed at meaningful, for odor perception, frequencies. In addition, as our spatial transcriptomic data showed, zonal regulation assures that ORs are expressed in a reproducibly patterned fashion in the MOE. While in wild type mice unbiased machine learning approaches identify at least 5 distinct OR expression patterns, in the triple Nfi cKO mice these patterns become intermixed for all but zone 1 ORs. With recent observations arguing that individual mitral cells, the second order neurons in the olfactory circuit, have patterned projections in the brain[49], non-random OR expression in the MOE may contribute to putative hardwired components of odor perception and valance[50].

Polygenic OR transcription as the arbiter between OR gene silencing and OR gene choice

A peculiar feature of the OR gene family that had emerged from our past work is that OR gene silencing is highest in the very cells that express ORs[20]. Our zonal analysis further strengthened this intriguing correlation, as both H3K9me3/H3K79me3 and genomic compartmentalization in each MOE segment are strongest on OR groups that are transcriptionally active during OSN differentiation. A fascinating implication from this observation is that early OR transcription is the signal for both genomic silencing and singular choice. Although the former is only implied by the strong correlation between OR transcription in OSN progenitors and genomic silencing, the latter is experimentally supported by the striking observation that strong transcriptional induction of Olfr17 at the INP/iOSN stage results in strong recruitment of the Greek Island hub, and stable choice of this OR allele in most mOSNs throughout the MOE. Such a mechanism of promoter choice influenced by spatially-determined early transcription, could also explain the recent demonstration that clustered Pcdh choice, which is regulated by anti-sense transcription[51], abides to spatial restrictions in the neocortex[6].

How could two fundamentally opposite gene expression outcomes be encoded on the same molecular feature? We propose that the timing and levels of transcriptional induction could be the arbiters between genomic silencing and singular choice. ORs that are transcribed first in the INP stage, when the Greek Island hub cannot yet form due to the continuous expression of Lamin b receptor[21], are most likely to be silenced. OR alleles activated during the assembly of the multi-enhancer hub, at the INP to iOSN transition, may compete for hub recruitment. The OR allele that will first associate with a multi-enhancer hub, will be stably protected from heterochromatic silencing, possibly due to the significantly increased rates of OR transcription, whereas the other co-transcribed ORs will succumb to heterochromatic silencing. If timing and rates of OR transcription determine whether an OR allele will be silenced or chosen, then an OR allele that is highly transcribed in both INP and iOSN stage should evade silencing and dominate the competition for hub recruitment, explaining the striking expression pattern of the tTA-induced Olfr17 allele. Thus, according to this model, in each OSN ORs with more dorsal identity will be transcribed first, because they have stronger promoters, and therefore will become silenced in higher frequency; ORs with the correct zonal identity will be transcribed later, with a chance to associate with the Greek Island hub, explaining why one is chosen and the rest are silenced; ORs with more ventral identities will not be transcribed at all, thus, will not be silenced but also will not be chosen. In other words, singular OR transcription may not depend on the silencing of every single OR in the genome: by encoding silencing and stable choice with the same exact molecular feature, OSNs choose one and silence a small fraction of the whole OR repertoire in each nucleus-the rest are not relevant.

Limitations of this study

Our experiments did not clarify whether NFI proteins bind directly on OR promoters, or act indirectly by activating other transcription factors and chromatin modifying enzymes. Although there is a statistically significant enrichment of NFI motifs on zone 4/5 OR promoters compared to the other OR promoters (data not shown), we were not able to detect direct binding of NFI proteins on these promoters, which is expected since these promoters are active in less than 1% of the cells. Given that our studies provide the mechanism by which NFI gradients establish zonal boundaries, via polygenic OR transcription and chromatin-mediated silencing, answering this question is not essential for understanding the mechanism of dorsoventral patterning of OR expression. A second limitation of this study is that it did not reveal the mechanisms that regulate the expression of the dorsal-most ORs (Zone 1 ORs), as NFI deletion had not effects in the expression and chromatin regulation of these OR genes. However, having revealed the regulatory logic whereby these patterns are established, we expect that other transcription factors with zonal expression patterns identified here, regulate early transcription and silencing of these genes across the MOE.

eLife Assessment

This is an important paper that revises the canonical model of how olfactory sensory neurons choose which odor receptor to express. The data presented in the paper are convincing and the model proposed is provocative and likely to enable important future work.

Reviewer #1 (Public Review):

Mature mammalian olfactory sensory neurons (OSN) express only one of the hundreds of possible odor receptors (ORs) encoded in the genome. The process of selecting this OR in each OSN is the consequence of both deterministic developmental processes involving transcription factors, and more stochastic processes. How this balance is implemented is a major problem in molecular neuroscience, one whose solution has significant systems-level implications for odor coding. In Bashkirova et al the authors substantially revise the canonical view of how this process works. By querying single cell transcriptomes and genetic architecture across OSN development, the authors demonstrate that OSN progenitors express ORs for their zone and for more dortsal zones, and that the degree of heterochromatinization of non-expressed ORs varies as a function of which zone a given OSN resides in. Through additional genetic experiments (including knockouts of transcription factors that seem to be associated with zonal identity, and the clever use of OR transgenes) they synthesize these findings into a model in which progenitors co-express many ORs - both ORs that are appropriate for their zone and ORs that are dorsal to their zone - and that this expression both facilitates heterochromatinzation of non-selected and extra-zonal ORs, and enables singular OR selection. The experiments are careful and the data are novel, and definitely revise our simplistic current view of how this process works; as such this work will have significant impact on the field. As presented the model requires additional experiments to fully flesh it out, and to definitively demonstrate that i.e., precocious expression leads to gene silencing, but with some additional clarifications in the discussion this paper both breaks new ground and sets the stage for future work exploring mechanisms of OSN development and OR selection.

Reviewer #2 (Public Review):

In this study, Bashkirova et al. analyzed how the gene choice of olfactory receptors (ORs) is regulated in olfactory sensory neurons (OSNs) during development. In the mouse olfactory system, there are more than 1000 functional OR genes and several hundred pseudogenes. It is well-established that each individual OSN expresses only one functional OR gene in a mono-allelic manner. This is referred to as the one neuron - one receptor rule. It is also known that OR gene choice is not entirely stochastic but restricted to a particular area or zone in the olfactory epithelium (OE) along the dorsoventral axis. It is interesting to study how this stochastic but biased gene-choice is regulated during OSN development, narrowing down the number of OR genes to be chosen to eventually achieve the monogenic OR expression in OSNs.

In the present study, the authors cell-sorted OSNs into three groups; immediate neuronal precursors (INPs), immature OSNs (iOSNs), and mature OSNs (mOSNs). They found that OR gene choice is differentially regulated positively by transcription factors in INPs and negatively by heterochromatin-mediated OR gene silencing in iOSNs. The authors propose that by the combination of two opposing forces of polygenic transcription (positive) and genomic silencing (negative), each OSN finally expresses only one OR gene out of over 2000 alleles in a stochastic but stereotypic manner.

The authors' model of OR gene choice is supported by well-designed experiments and by large amounts of data. In general, the paper is clearly written and easy to follow. It will attract a wide variety of readers in the fields of neuroscience, developmental biology, and immunology. The present finding will give new insight into our understanding of gene choice in the multigene family in the mammalian brain and shed light on the long-standing question of monogenic expression of OR genes.

Reviewer #3 (Public Review):

This manuscript investigates how a seemingly random choice of odourant receptor (OR) gene expression is organised into sterotypic zones of OR expression along the olfactory epithelium. Using a varietty of functional genomics methods, the authors find that along the differentiation axis (progenitor to mature olfactory sensory neuron, OSN) multiple ORs are initally transcribed and from among these, only one OR is selected for expression. The rest are suppressed through chromatin silencing. In addition to this, the authors report a dorso-ventral gradient in OR expression at the immature stage - dorsally expressed ORs are also expressed ventrally and these then get silenced. The expression of the ventrally expressed ORs, on the other hand, are restricted to the ventral region. They suggest a role for the transcription factor NF1 in this dorsoventral process.

This is a valuable study. The data are compelling and generally well presented.

Author Response

The following is the authors’ response to the original reviews.

We are grateful to the 3 reviewers and the editorial team for agreeing that our work is rigorous and valuable for the fields of olfaction and developmental biology. We provide a revised version of the manuscript that addresses major concerns raised by the reviewers and adheres to their suggestions.

Specifically:

-We clarify what is novel in this work and we cover the appropriate literature.

-We tone down the language and interpretation of our data

-We clarify the categorization of zones and improve the readability to the best of our ability.

We have also made every effort to address minor points raised by the 3 reviewers and made clarifications wherever requested.

Version 2

Abstract

Abstract

Olfactory receptor (OR) choice represents an example of genetically hardwired stochasticity, where every olfactory neuron expresses one out of ∼2000 OR alleles in a probabilistic, yet stereotypic fashion. Here, we propose that topographic restrictions in OR expression are established in neuronal progenitors by two opposing forces: polygenic transcription and genomic silencing, both of which are influenced by dorsoventral gradients of transcription factors NFIA, B, and X. Polygenic transcription of OR genes may define spatially constrained OR repertoires, among which one OR allele is selected for singular expression later in development. Heterochromatin assembly and genomic compartmentalization of OR alleles also vary across the axes of the olfactory epithelium and may preferentially eliminate ectopically expressed ORs with more dorsal expression destinations from this “privileged” repertoire. Our experiments identify early transcription as a potential “epigenetic” contributor to future developmental patterning and reveal how two spatially responsive probabilistic processes may act in concert to establish deterministic, precise, and reproducible territories of stochastic gene expression.

Introduction

The development of multicellular organisms relies on gene expression programs that are precisely regulated in space and time. To transform probabilistic biochemical reactions, such as transcription and translation, into reproducible differentiation processes, plants and animals convert individual cellular variability into predictable cell population averages. Yet, there are cases in biology where gene expression variability is desirable, as it generates diverse cellular identities that are difficult to obtain with deterministic gene regulation. For example, production of antibodies via VDJ recombination, and evasion of immunological responses by antigenic variation represent biological systems that seek utmost randomness[1, 2]. Other biological functions, however, benefit from balancing absolute determinism with complete randomness, producing biased stochasticity. Genetically encoded biased stochasticity is often observed in the nervous system, where gene expression choices generated by neurons must integrate into functional and reproducible circuits[3]. In fly ommatidia, for example, biased randomness preserves a ratio of photoreceptor neuron identities across animals[4], whereas in mammals, random Protocadherin promoter choice[5], was recently shown to obey spatial patterns in the mouse neocortex, assuring proper tiling between neighboring neurons[6].

Mammalian olfactory receptor (OR) gene choice provides an extreme case of hardwired biased randomness[7]. OR transcription starts in neuronal progenitors of the main olfactory epithelium (MOE), which transiently express 5-15 ORs out of >1,000 OR genes distributed in genomic clusters across chromosomes[8-10]. As these progenitor cells differentiate into post-mitotic olfactory sensory neurons (OSNs), they switch from polygenic to monogenic and monoallelic OR transcription[11]. This transition is mediated by the assembly of a multi-chromosomal enhancer hub over the chosen OR allele[12-14], followed by the stabilizing effects of an OR-elicited feedback signal[15-19]. During this developmental progression, heterochromatic silencing[20] and genomic OR compartmentalization[21, 22] act together to assure that the non-chosen OR alleles will remain transcriptionally repressed for the life of the OSN. Interestingly, the position of the OSN across the dorsoventral (DV) axis of the MOE predisposes this singular transcriptional choice towards a group of 50-250 OR genes[23], providing reproducible topographic restrictions in OR expression. The anatomical segments of the MOE that express a specific collection of OR identities are known as “zones”, with their total number varying from 4 to 9, depending on the analyses and criteria used to define them[24-27]. Although zonal restrictions in OR expression have a well-established influence in the wiring of the olfactory circuit[28, 29], the mechanisms that bias this singular transcriptional choice towards specific OR identities remain unknown.

Here, we identified genetically encoded mechanisms that introduce topographic biases in OR gene regulation. We report that OSN progenitor cells from various MOE segments transcribe OR mixtures consisting of ORs with the corresponding or with more dorsal expression identities. Ectopic expression of dorsal identity ORs at the polygenic stage of OR transcription is rectified during differentiation by preferential genomic silencing that is skewed towards ORs with more dorsal expression identities than the identity of the OSN. Patterns of polygenic OR transcription and genomic OR silencing are influenced by gradients of transcription factors NFI A, B, and X [30]. Triple NFI (NFIA, B and X) deletion both eliminates heterochromatic silencing and genomic compartmentalization from ORs with ventral identities, as well as extinguishes their transcription in olfactory progenitors. Furthermore, spatial transcriptomics revealed a dramatic expansion of dorsomedial identity OR expression towards the ventral MOE and reciprocal transcriptional reduction of ventral identity ORs in triple NFI knockouts (cKOs), suggesting that patterns of genomic OR silencing and polygenic OR transcription influence OR gene choice. Indeed, transcriptional induction of an OR allele in OSN progenitors biases the choice towards this allele in mature OSNs (mOSNs) throughout the MOE. Strikingly, by modulating the levels of OR induction in progenitor cells we can restrict expression of this OR allele to more dorsal OSNs, where heterochromatic silencing and genomic compartmentalization is less prevalent. Thus, our studies reveal that position-responsive OR transcription in OSN progenitors may act as an “epigenetic” signal for future singular choice among the previously transcribed ORs. Moreover, our data suggest that polygenic OR transcription and heterochromatic silencing/genomic compartmentalization could act as opposing regulatory “rheostats” that determine in a spatially influenced fashion the exact OR repertoire that is available for stochastic singular choice in mature OSNs.

Results

OSN progenitors co-transcribe an increasing number of zonal OR identities towards the ventral MOE

The mouse MOE is divided into a limited series of stereotypic segments, based on expression of OR genes, that exhibit bilateral symmetry between the two nasal cavities (Figure 1A). In whole mount views, these segments present a dorsoventral (DV) segmentation pattern, with zone 1 being at the dorsal and zone 5 at the ventral end of the MOE. Intricate invaginations of the MOE occurring during embryonic development and early postnatal growth convolute this dorsoventral segmentation pattern, especially when viewing coronal sections of the MOE (Figure 1A). However, we will continue referring to the DV coordinates of each one of the five segments, or zones, as they correspond to their initial patterning during development.

Polygenic transcription of OR genes in olfactory progenitors follows a zonal expression pattern.

(A) Schematic illustrating OR zones along the dorsoventral axis in whole mount views of the MOE (left) and coronal sections (middle). Zone1 (red) is the dorsal-most zone and zone 5 (blue) is the ventral-most zone. Zoomed in view of the MOE (right) shows cell populations at different stages of OSN differentiation organized in a pseudostratified fashion from the basal (least differentiated) to apical (most differentiated) layers: HBC, horizontal basal cell; GBC, globose basal cell; INP, immediate neuronal precursor; iOSN, immature olfactory sensory neuron; mOSN, mature olfactory sensory neuron. MOE: main olfactory epithelium, OB: olfactory bulb.

(B) t-SNE (t-distributed Stochastic Neighborhood Embedding) dimensionality reduction used to visualize clustering of single cells from FAC-sorted MOE cell populations with Seurat, based on expression of the most variable genes. Plot (left panel) shows the separation of single cells into 6 populations, to which we assigned cell identities based on expression of known MOE markers [31] (See also Supplementary Figure S1). Olfactory receptor expression is first detected in INP3 cells (right panel).

(C, D, left panel) t-SNE plot (as in B) of cell populations isolated from either dorsal (zones 1-2) in (C) or ventral (zones 3-5) in (D) MOE microdissections. Cells are colored according to the zonal identity of the most highly expressed OR. (C, D, right panel) Plots depicting zonal identities of all the OR genes detected in individual INP3 cells from dorsal (C) or ventral (D) MOE. Y axis shows OR expression in normalized counts of unique transcripts (UMIs) for different OR genes (separated by black lines). On the X axis each point is a different INP3 cell. ORs are colored according to their zonal identity. Note that while class I OR genes are expressed within zone 1 of the MOE, they may be regulated through a different mechanism and are thus displayed separately.

(E) Expression of OR genes of different zonal identities in olfactory progenitor INP cells (top) and mOSNs (bottom), determined with bulk RNA-seq, in cells isolated from dorsal-most (zone 1) (left), dorsomedial (zones 2-3) (middle), and ventral-most (zones 4-5) MOE microdissections (right). Note that INP and mOSN cells were FAC-sorted from the same exact dissection, thus the mOSN OR expression patterns confirm the accuracy of the dissection.

Within each zone, mOSNs express a single OR allele among 50-250 OR genes with proper zonal identities. However, before the onset of singular OR expression, mitotically active OSN progenitors, the immediate neuronal precursor (INP) cells, co-express multiple lowly expressed ORs [8-10]. To determine whether zonal restrictions are operational from this polygenic stage of OR transcription, we performed plate-based single cell RNA-seq (scRNA-seq) analysis of FAC-sorted OSNs and OSN progenitors isolated from the MOE, which was micro-dissected into two parts: a more dorsal (zone 1-2) and a more ventral (zone 3-5) segment (Supplementary Figure S1A). With a median of over 130,000 unique transcripts per cell, this method readily detected low level OR gene transcription in OSN progenitors. Additionally, to enrich our plate-based scRNA-seq for cell populations of interest we used Mash1-CreER; tdTomato; Ngn1-GFP triple transgenic mice (Supplementary Figure S1A). We injected P2 mice with tamoxifen, inducing permanent tdTomato expression, and then collected cells 48 hours later (Supplementary Figure S1A). From each dissection we isolated four major cellular populations corresponding to four successive differentiation stages, as previously described [31, 32]: GBCs (MOE stem cells), INPs (immediate neuronal precursors), iOSNs (immature OSNs), and mOSNs (Supplementary Figure S1A-B).

Dimensionality reduction and clustering of single cell RNA-seq data with Seurat[33] sorted cells into 6 populations. We determined each population’s stage of OSN development using expression of known MOE markers (Figure 1B, Supplementary Figure S1B-C). We first detect OR mRNAs in INP3 cells (Figure 1B), which consistently transcribe multiple ORs. Next, we examined the zonal identity of ORs co-transcribed within individual dorsal or ventral INP3 cells. We created a separate category for the ∼100 “class I” OR genes, which were grouped based on homology and resemblance to OR genes first identified in fish. Virtually all class I OR genes are expressed in zone 1, but their expression is likely regulated though a separate mechanism [34, 35], thus we chose to analyze them separately. Surprisingly, while dorsally positioned INPs transcribe almost exclusively dorsal identity ORs, ventrally positioned INPs transcribe complex mixtures consisting of ORs of every zonal identity (Figure 1C-D). Overall, ventral INPs transcribe a greater number of OR genes compared to dorsal INPs. Focusing on OR genes detected with at least 3 unique transcripts, we detect dorsal identity ORs in 43 ventral INP cells and ventral identity ORs in only 29 of them, while dorsal INP cells express predominantly dorsal ORs (Figure 1C-D). Moreover, as the ventral INPs differentiate to iOSNs, dorsal identity OR transcription is replaced by the “correct” (zone-appropriate) ventral OR transcription, culminating in singular expression of an OR allele with the correct zonal identity in mOSNs (Figure 1D). These observations were independently confirmed by bulk RNA-seq on FAC-sorted INP and mOSN cells extracted from trisected dorsal (zone 1), dorsomedial (zone 2-3), and ventral (zone 4-5) MOE, using the same labeling and FAC-sorting approach used for the single cell experiments. This bulk analysis showed that in every case INPs co-transcribe ORs with the correct as well as more dorsal zonal identities, while further differentiation replaces dorsal ORs with ORs of the correct identity (Figure 1E). This finding immediately poses mechanistic questions about the process that shuts off dorsal ORs and enhances the transcription of the ORs expected to be expressed in each MOE segment.

Heterochromatin eliminates ectopically expressed ORs along the dorsoventral MOE axis

We previously showed that OSN differentiation coincides with heterochromatin-mediated OR gene silencing [20]. If heterochromatinization contributes to singular OR choice by eliminating every non-chosen OR allele transcribed in INPs, then in any MOE segment silencing should be preferentially applied to ORs with the correct or more dorsal zonal identities. We performed native ChIP-seq in the MOE to determine deposition of histone marks associated with heterochromatin including H3K9me3, a marker of constitutive heterochromatin, and H3K79me3, which we also find labeling heterochromatin on OR gene clusters[13, 36]. We predicted that dorsal-most identity (zone 1) ORs, which are expressed in INPs throughout the MOE, should have the highest levels of heterochromatin, whereas ventral-most identity (zone 5) ORs, which are transcribed only in ventral INPs, should have the lowest, with the rest of the OR repertoire having intermediate levels of heterochromatin marks. Indeed, visual inspection of ChIP-seq genomic tracks along OR gene clusters with mixed zonal constitution reveals highest H3K9me3/H3K79me3 levels on the dorsal identity OR genes and lowest on the ventral identity OR genes of the cluster (Figure 2A). Aggregate ChIP-seq analysis of all the OR genes grouped by their zonal identities corroborates the gradual reduction of H3K9me3 and H3K79me3 enrichment from dorsal to ventral ORs for the whole OR repertoire (Figure 2B). The only exception from this pattern is found at the dorsally expressed class I ORs, which rely on different regulatory mechanisms than the canonical class II ORs [34, 35] (Figure 2B). Finally, using the FACS-based strategy described earlier, we confirmed that both heterochromatic marks are predominantly deposited during the INP to iOSN transition, simultaneously with the transition from polygenic to singular, zonally appropriate OR expression (Figure 2C, Supplementary Figure S2A). Importantly, the descending pattern of heterochromatin enrichment from dorsal to ventral OR identities is preserved throughout differentiation.

Heterochromatin deposition silences OR genes from lower zones.

(A) Signal tracks of H3K9me3 and H3K79me3 native ChIP-seq from the whole MOE show heterochromatin deposition over two representative OR gene clusters. These clusters were selected because they harbor OR genes with both dorsal (zone 1) and ventral (zone 5) identities. OR genes are colored according to their zonal identity: zone 1 ORs in red, zone 2 ORs in yellow, zone 5 ORs in blue and ORs with unknown zonal identity in gray. Purple triangle marks the “H” OR gene enhancer that is present within that OR gene cluster.

(B) H3K9me3 (left) and H3K79me3 (right) native ChIP-seq in the whole MOE. Box plots of read density over OR gene bodies, separated by their zonal identity, depict a pattern of deposition that is high on dorsal-most (zone 1) OR genes, progressively decreases with more ventral zonal OR identities, and is absent on class I ORs.

(C) H3K79me3 native ChIP seq in GBC, INP, iOSN and mOSN populations shows onset of H3K79me3 deposition as cells transition from INPs to iOSNs. Each row of the heatmaps shows coverage over an OR gene body (separated into categories by their zonal identity). (See also Supplementary Figure S2A for H3K9me3 heatmap).

(D) H3K79me3 native ChIP-seq in mOSNs from zonally dissected MOE. Colored schematics above each heatmap depict the zone of dissection. (See also Supplementary Figure S2B for H3K9me3 heatmap).

(A-D) Pooled data from two biological replicates is shown for all ChIP experiments.

We then asked if the patterns of heterochromatin deposition detected in mixed OSNs from the whole MOE are preserved in distinct zones. We performed ChIP-seq in mOSNs isolated from dissected dorsal (zone 1), dorsomedial (zone 2-3) and ventral (zone 4-5) segments of the MOE. In each segment, OR genes with either the correct or more dorsal zonal identities had a higher level of heterochromatin compared to more ventral identity ORs (Figure 2D, Supplementary Figure S2B). Intriguingly, this is the same zonal pattern observed for OR gene transcription in INP cells. Only the OR gene identities able to be transcribed in the INP cells of a given zonal MOE segment acquire heterochromatin. Thus, most OR genes are heterochromatic in ventral OSNs; dorsal and dorsomedial identity OR genes are heterochromatic in dorsomedial OSNs; and only dorsal identity OR genes have some heterochromatin in dorsal OSNs (Figure 2D, Supplementary Figure S2B). Although each zonal identity OR group is heterochromatic in the MOE segment where it is expressed by mOSNs, the level of H3K9me3/H3K79me3 enrichment is lower than in more ventral segments, where it is not chosen for stable expression. Thus, dorsal identity ORs have less heterochromatin in dorsal OSNs than in the rest of the MOE, and dorsomedial identity ORs have less heterochromatin in dorsomedial OSNs than ventral OSNs. Similarly, at the ventral end of the DV axis, ventral identity ORs have less heterochromatin than dorsal and dorsomedial identity ORs. Detection of heterochromatin on OR genes with the correct zonal identity is not counterintuitive, as only one OR allele from the ones co-transcribed will be eventually chosen, and the rest must be silenced. Thus, in a pure population of ventral mOSNs expressing Olfr1507 (a ventral, zone 5 identity OR), the remaining non-chosen zone 5 identity OR genes are silenced with the same level of heterochromatin as OR genes with more dorsal zonal identities (Supplementary Figure S2C). In other words, in every MOE segment, OR heterochromatinization is preserved only for the ORs that have the potential to be expressed and is not applied to more ventral ORs, which were not transcriptionally active in INPs. This is consistent with recent reports of heterochromatin marks being detected on trace amine-associated receptor (TAAR) genes only in TAAR-expressing OSNs and not the rest of the MOE[37].

DV gradient of OR gene compartmentalization follows patterns of heterochromatin assembly

Heterochromatic OR genes converge into multi-chromosomal genomic aggregates of extreme chromatin compaction that contributes to the effective and stable silencing of non-chosen ORs [21]. We thus asked if the spatially determined pattern of OR heterochromatinization at the linear genome coincides with a similar pattern of 3D genomic compartmentalization. We performed in situ Hi-C in FAC-sorted mOSNs isolated from MOE segments along the DV axis and determined the frequency with which OR genes form long range interactions in each segment. We saw a striking resemblance between deposition of heterochromatic marks on OR genes and genomic compartmentalization (Figure 3A-B, Supplementary Figure S2D). For example, inspection of the long range cis genomic interactions between 3 OR gene clusters on chromosome 2 shows that a cluster of predominately ventral identity OR genes is recruited to OR compartments only in ventral OSNs, where they are heterochromatic (Figure 3A). In contrast, the other two OR gene clusters, which are either enriched for dorsal ORs, or have mixed constitution, make strong genomic contacts with each other in all three MOE segments (Figure 3A). To expand this analysis to every OR, we measured the frequency of interchromosomal trans genomic interactions between OR genes with different DV identities. To do so, we annotated each OR gene cluster bin at 50kb resolution according to the zonal identity of the residing ORs and plotted the average interchromosomal contacts between bins of different constitution, excluding class I OR genes, which formed very few contacts with other ORs. This analysis yielded the same conclusion: Interactions between dorsal OR genes is the default in every OSN, whereas compartmentalization for the remaining of the OR repertoire increases along the DV axis of the MOE (Figure 3B). Intriguingly, as with levels of heterochromatin, we detect the following recurrent pattern of OR compartmentalization: every OR has intermediate Hi-C contact frequencies with other OR genes in their segment of expression in the MOE, lower Hi-C contact frequencies in more dorsal MOE segments, and higher Hi-C contacts in more ventral segments.

Zonal OR compartmentalization permits OR genes from more ventral zones to be recruited into the OR compartment.

(A) In situ Hi-C contact matrices of a 90Mb region of chromosome 2 that contains 3 large OR gene clusters, depicted with boxes under the contact matrices. Hi-C libraries were prepared from mOSNs FAC-sorted from dorsal-most (zone1) and ventral-most (zone 4-5) MOE microdissections, as well as a pure population of Olfr17 (a zone 2 OR) expressing dorsomedial mOSNs. For each zonal contact matrix, magnified views show long-range cis Hi-C contacts between the large OR gene cluster in the middle that contains ORs of every zonal identity with the OR cluster on the left that contains mostly zone 1-2 identity ORs (red box) and the OR cluster on the right that contains mostly zone 4-5 identity ORs (blue box). Cis contacts between OR genes increase from dorsal to ventral mOSNs, but the zone 4-5 identity OR cluster associates with the other ORs only in the ventral-most OSNs (as seen when comparing Hi-C contacts in the blue boxes).

(B) Heatmaps of average interchromosomal Hi-C contacts between OR genes annotated by their zonal identity at 50Kb resolution show increased trans contacts in mOSNs from more ventral zones. OR genes have a similar, intermediate frequency of contacts in the mOSN population where they are expressed, marked with an asterisk. Class I OR genes (which are also expressed in zone 1) make few interchromosomal interactions in all zones (data not shown) and were thus excluded from this analysis.

(C) Dip-C in mOSNs from dorsal and ventral dissected MOE was used to generate haplotype resolved single cell contact matrices and 3D genome structures, as previously described [40].

(D) Analysis of Dip-C contact densities of interchromosomal contacts between ORs genes confirms that ventral mOSNs have increased OR compartment interactions (Wilcoxon rank sum p-value = 9.164e-11).

(E) Single cell heatmaps of pairwise distances between OR genes generated from 3D genome structures in two ventral mOSNs show OR genes from different chromosomes intermingle in a different pattern in the two cells (top). For each cell, heatmaps are sorted by chromosome order and show all OR interactions within 10 particle radii (approximately ∼600nm). Representative 3D structures show the different positioning of three chromosomes (chr19, chr9 and chr2) in the two cells, resulting in a different pattern of OR cluster contacts (bottom). See also Supplementary Figure S3 for heatmaps of Dip-C distances in each of the 48 dorsal and ventral mOSNs.

The “intermediate” levels of heterochromatin enrichment and Hi-C contacts observed on OR genes within their segment of expression in the MOE may reflect a less compact, transcription-compatible state of heterochromatin, or less frequent silencing of these OR genes compared to more dorsal identity ORs. To distinguish between the two scenarios, we explored OR silencing at the single cell level using Dip-C, a single cell Hi-C method [38-40]. We performed Dip-C in 48 dorsal and 48 ventral mOSNs (Figure 3C). To retain allelic information, we used heterozygous mice from a cross between OMP-IRES-GFP (a mOSN reporter) and Castaneous (Cas) mice, and used Cas-specific SNPs to distinguish Cas from non-Cas alleles. Analyzing single cell genomic contacts, we saw a greater enrichment of contacts between OR gene loci in ventral cells relative to dorsal cells, consistent with our bulk Hi-C data (Figure 3D). We then used the haplotype resolved chromatin contacts to compute distances of all genomic loci at 20kb resolution, from which we generated 3D genome structures for all cells (Figure 3C), as previously described [40]. Analyzing distances between pairs of OR loci in the 3D genome structures we determined the size and complexity of OR compartments in each cell. We confirmed that OR compartments in ventral mOSNs are larger and contain more OR genes from more chromosomes than in dorsal mOSNs (Supplementary Figure 3C). Importantly, in each cell, significantly fewer ventral identity OR genes were found in OR compartments compared to dorsal (zone 1) or dorsomedial (zone 2-3) identity OR genes (Supplementary Figure 3D). From this result we can conclude that the “stronger” Hi-C contacts among dorsal ORs observed in bulk, represent an increased number of dorsal ORs participating in OR compartments in each OSN. Thus, extrapolating Dip-C results to H3K9me3/H3K79me3 enrichment, we conclude that “intermediate” silencing levels of each OR group in their own zone likely reflect less frequent silencing of these ORs than ORs with more dorsal zonal identities. In this note, OR compartmentalization is highly probabilistic, with each one of the 48 dorsal and ventral OSNs having unique maps of OR-OR genomic interactions (Figure 3E, Supplementary Figure 3A, B). Thus, we propose that the balance between two probabilistic yet DV-responsive processes, early transcription and genomic silencing, may define the OR ensemble that is available for singular choice along the DV axis. To test this model, we sought to identify factors responsible for generating these remarkable patterns.

NFI paralogues generate DV patterns in OR expression

We searched our RNA-seq data for transcription factors that have strong expression during the INP to iOSN transition that is graded across the DV axis of the MOE. NFI paralogues NFIA, B, and X have strong, DV-influenced expression in INPs that is preserved in iOSNs (Figure 4A, B and Supplementary Table 1). Specifically, NFIA and NFIB are expressed higher in ventral INPs and iOSNs, and NFIX is higher in ventral mOSNs (Figure 4B). These three members of the nuclear factor I (NFI) family of transcription factors control a plethora of developmental and cell specification processes [30, 41], and were previously implicated in OSN differentiation [42, 43]. Thus, we decided to genetically explore their contribution in the establishment of dorsoventral patterns of OR expression.

NFI paralogue gradients regulate zonal OR expression

(A) Heatmaps showing differentially expressed transcription factors in immediate neuronal precursor (INP) and immature olfactory sensory neuron (iOSN) cells isolated from either dorsal, dorsomedial or ventral olfactory epithelium. The shown transcription factors are significantly differentially expressed between dorsal and ventral cells with an adjusted p-value of <0.05, at least a three-fold change in expression, and an expression level of at least 15 TPM. A broader list of zonal transcription factors is included in Supplementary Table S1. The heatmaps are sorted based on expression in ventral cells and the color bar above each heatmap shows the log2 fold change in ventral cells relative to dorsal cells.

(B) Expression levels of NFIA, NFIB, and NFIX at four stages of OSN development in dorsal cells (red), dorsomedial cells (green) and ventral cells (blue).

(C, D) Comparison of OR gene expression in NFI ABX triple knockout and control cells from the whole MOE. NFI transcription factors are deleted either in olfactory progenitors (C) using the Krt5-CreER driver or in mOSNs (D) using the OMP-IRES-Cre driver (as illustrated in Supplementary Figure S4). At the right of each panel, scissors indicate the differentiation stage of NFI ABX deletion, and a red box marks the cell type that was FAC-sorted for RNA-seq analysis. Asterisk: Wilcoxon rank sum test p-value < 0.001.

(E) OR expression in NFI ABX triple knockout, NFI AB double knockout, NFIX knockout, NFI ABX triple heterozygous and control mOSNs from ventrally dissected MOE. Knockout was induced in progenitors with the Krt5-CreER driver. Plots show a different pattern of OR gene transcription in the different genotypes. Quantification of differentially expressed ORs for the three knockout genotypes is shown in Supplementary Figure S4.

To interrogate the potential role of NFIA, B and X in zonal OR expression we deleted all three genes simultaneously using the Krt5-CreER driver, which is expressed in the quiescent stem cells of the MOE (HBCs). We crossed Krt5-CreER; tdTomato mice to NFIA, B, X fl/fl mice [44], and induced recombination with tamoxifen. To force the quiescent HBCs to differentiate into OSNs, we ablated the MOE with methimazole and allowed 40 days for a complete restoration by the marked progeny of the NFI triple conditional knockout (cKO) or control HBCs (Supplementary Figure S4A-B), as previously described [14]. RNA-seq analysis of the FAC-sorted cKO OSNs from the whole MOE revealed significant transcriptional reduction of ventral OR identities and reciprocal increase of dorsomedial ORs (Figure 4C). In contrast, triple NFI deletion only in mOSNs, using the OMP-IRES-Cre driver, has no measurable effects on OR expression (Figure 4D). To determine whether the reduced transcription of ventral ORs reflects a developmental defect of ventral OSN differentiation, versus a bona fide dorsalization of ventral OSNs, we performed RNA-seq in cKO OSNs isolated specifically from ventral MOE microdissections. This experiment revealed ectopic expression of OR genes with dorsomedial zonal identities in place of the OR genes with proper ventral ones (Figure 4E), a result confirmed by immunofluorescence (IF) experiments (Supplementary Figure S4D,E). This transcriptional transformation of ventral OSNs satisfies the original criteria of homeosis [45], since the overall mOSN identity is not altered by the triple NFI deletion: Only 13 out of ∼200 OSN-specific genes are significantly different between control and cKO OSNs, and 117/207 non-OR ventral markers are still expressed in the ventral-most zones, acting as independent fiducial markers for our zonal dissection (Supplementary Figure S4F). Interestingly, deletion of just one NFIA, B, and X allele had an intermediate effect, with expression of ventral-most (zone 5) ORs replaced with zone 4 ORs in ventral mOSNs (Figure 4E). In fact, the severity of this dorsomedial transformation depends on the total number of NFI alleles deleted, with the triple NFIA, B, X cKO mOSNs expressing predominantly zone 2 and 3 ORs, double NFIA, B cKO mOSNs expressing zone 3 and 4 ORs, and single NFIX cKO mOSNs having almost wild type expression patterns of zone 4 and 5 ORs (Figure 4E, Supplementary Figure S4C).

Spatial transcriptomics reveals widespread homogenization and dorsalization of the MOE upon triple NFI deletion

To obtain a complete and unbiased understanding of the consequences of triple NFI deletion on patterns of OR expression, we deployed a spatial transcriptomic approach. Since our goal was to decipher zonal patterns of OR expression across the dorsoventral MOE axis without requirements for single cell resolution, we opted for the Visium Spatial Gene Expression workflow (10X Genomics) [46], rather than the previously described TOMO-seq approach[47]. This workflow is ideal for interrogation of spatial OR expression in mOSNs, as OR mRNAs are highly abundant and readily detectable in most spatial spots that contain OSN mRNAs. For increased stringency, we only included spatial spots that include more than 2 OR genes and 3 OR transcripts. We analyzed 2 MOE sections each from an NFI ABX triple cKO and age-matched control mouse (Figure 5A). Expression data on OR genes were normalized and integrated across replicates (see methods). We performed PCA analysis, by which spatial spots were arranged in 5 clusters in control and cKO MOEs (Figure 5B). Interestingly, while dimensionality reduction and unbiased clustering generated OR clusters that correspond to zonal patterns of OR expression (i.e., each cluster contains OR genes with one zonal identity) in control MOEs, only dorsal-most zone 1/class I OR genes followed this correlation in cKO MOEs (Figure 5B). The other 4 clusters in the cKO homogenously express dorsomedial zone 2-4 ORs, with expanded expression of zone 2 ORs in every cluster and complete loss of ventral-most zone 5 OR expression. Thus, conditional triple NFI deletion causes loss of spatial patterning for zone 2-4 OR genes and loss of expression for zone 5 OR genes, without influencing the expression of zone 1 ORs.

Spatial transcriptomics shows dorsalization and homogenization of the MOE upon NFI A, B, and X deletion.

(A) Schematic depicting our analysis pipeline: Spatial transcriptomics was performed on sections of wt control and NFI ABX cKO MOE. Dimensionality reduction was performed, and spatial spots were clustered based on normalized expression of OR genes.

(B) Heatmaps showing scaled, normalized expression levels of the top 20 highest expressed OR genes per zone in the control dataset. Unbiased neighborhood analysis and clustering grouped spatial spots into 5 clusters for both control and cKO MOE (depicted in distinct colors on the top of the heatmaps). Clustering of spatial spots in the control sample reproduces anatomical zones, as spots within each cluster express OR genes with the corresponding zonal identity (left heatmap). The same clusters were generated for NFI cKO sample (right heatmap). Although cluster 1 expresses exclusively zone 1 ORs, like in control MOEs, clusters 2-5 exhibit homogenous OR expression, with ventral expansion of zone 2/3 ORs, and reduced representation of zone 4/5 ORs.

(C) Average normalized per-spot expression of the 20 highest expressed OR genes from zone 1, zone 2, and zone 5 is overlaid against H&E histological image of control (top) and NFI cKO (bottom) MOE sections. Expression of zone 1 OR genes is confined to the same anatomical region for both control and NFI cKO sections. Zone 2 OR gene expression is spread to more ventral regions in the NFI cKO compared to control sections, while expression of zone 5 OR genes is almost completely absent in the NFI cKO sample.

(D) Spatial spots are colored according to their zonal assignment, which was determined based on the highest summed normalized expression of OR genes per zonal identity within that spot. Zonal spot assignment of the control sample visually reproduces known anatomical zones. In the NFI cKO sample, spots in the dorsal region have the highest expression of class I and zone 1 OR genes, similar to the control sample. However, in the rest of the NFI cKO MOE, most spots have a zone 2 OR identity. Spots assigned the identity ‘none’ did not contain any OR transcripts and were excluded from cluster analysis.

To depict the effects of triple NFI deletion on spatial patterns of OR expression, we plotted the average OR expression per spatial spot of the top 20 most highly expressed OR genes with dorsal-most (zone 1), dorsomedial (zone 2), and ventral-most (zone 5) identities. We then overlaid the corresponding values against the histological images of the control (wt) and NFI ABX cKO MOEs (Figure 5C). As observed in the clustering and heatmap analysis, dorsal-most zone1 OR expression is confined to the same anatomical region for both samples. However, dorsomedial zone 2 OR expression in the cKO MOE extends beyond its defined anatomical region from the control MOE, and spreads to the ventral-most zones (Figure 5C). This expansion is also observed in the expression of individual zone 2 OR genes (Supplementary Figure S5A). In contrast, the top 20 zone 5 OR genes, while highly expressed in control MOEs, are almost undetectable in NFI cKO MOEs (Figure 5C), consistent with our RNA-seq analysis. Olfr1507, the most highly expressed zone 5 OR, is undetectable in the cKO spatial spots (Supplementary Figure S5B), in agreement with our IF data. Finally, to obtain a more general understanding of the spatial transformations in OR expression patterning upon triple NFI deletion, we assigned a zonal identity to each spatial spot using the maximum normalized expression of all the OR genes detected within a spot (see methods). Unlike control MOEs, where spot assignment reproduces zonal anatomical positions, most spatial spots in the cKO MOEs, excluding the unchanged zone1, are assigned a zone 2 identity. Even the few spots assigned a zone 3 identity are shifted towards more ventral positions within the MOE relative to control, in a striking dorsalization and homogenization of the MOE (Figure 5D).

NFI gradients control patterns of OR heterochromatinization and polygenic OR transcription

We searched for a mechanistic explanation for the homeotic transformation of ventral OSNs in triple NFI cKO mice. Our experiments so far have identified 3 spatially responsive processes that may contribute to the dorsoventral patterning of OR gene choice: polygenic OR transcription in INPs, OR heterochromatinization and genomic compartmentalization during the INP to iOSN transition. Thus, we explored the effects of triple NFI deletion in all three processes. First, we investigated the effects of NFI deletion on OR heterochromatinization with ChIP-seq in triple NFI cKO OSNs from the ventral-most MOE segments. ChIP-seq revealed an almost complete loss of heterochromatin from ventral ORs as well as a reduction on zone 3 dorsomedial ORs in NFI cKO ventral OSNs and INPs (Figure 6A, Supplementary Figure S7F-G). Similarly, in situ Hi-C in control and triple NFI cKO OSNs from ventral MOE segments revealed a strong reduction in the long-range cis and trans genomic contacts made by ventral ORs (Figure 6B). In contrast, dorsal ORs did not exhibit strong changes in ChIP-seq and Hi-C contacts (Figure 6A, B). Strikingly, in both processes, heterochromatin assembly and genomic compartmentalization, the patterns observed in ventral OSNs upon NFI deletion are similar to those observed in dorsomedial OSNs from the control MOEs (Supplementary Figure S6A, B).

NFI A, B and X regulate chromatin state and OR compartment formation

(A) Native ChIP-seq for H3K9me3 (top) and H3K79me3 (bottom) in NFI ABX knockout mOSNs from ventral MOE. Heatmaps show ChIP signal over OR gene bodies, scaled to 6kb with 2kb flanking on either side. There is a decrease of both histone marks on zone 3-5 identity OR genes in NFI ABX knockout compared to control. Triple NFI deletion was induced with the Krt5-CreER driver (before OSN differentiation).

(B) Hi-C in in NFI ABX knockout and control mOSNs from ventral MOE. Left: In situ Hi-C contact matrices of a 90Mb region of chromosome 2 from control (top) and NFI ABX triple knockout (bottom) ventral mOSNs, as described in Figure 3A. The contact matrix shows long-range cis interactions between 3 large OR gene clusters: one enriched for dorsal, zone 1-2, identity ORs (left), one containing ORs of every zonal identity (middle), and one enriched for ventral, zone 4-5, identity ORs. Note that long range cis contacts between the zone 4-5 identity enriched cluster and the mixed identity cluster dissipate in the triple NFI cKO (bottom, blue box), whereas the contacts of the mixed identity cluster with the zone 1-2 identity enriched cluster are preserved (bottom, red box). Right: Heatmaps of average interchromosomal Hi-C contacts between OR genes annotated by their zonal identity (as described in Figure 3B) in control (top) and triple NFI cKO (bottom) mOSNs from ventral MOE. Trans contacts between zone 4-5 ORs dissipate, whereas trans contacts between zone 2-3 ORs reach intermediate levels typically detected between OR genes with the “correct” zonal identity for a given MOE segment (see Figure 3).

(C) OR expression by zonal identity in INP cells isolated from ventral NFI ABX knockout and control MOE. Triple NFI deletion was induced with the Krt5-CreER driver (before OSN differentiation) and NFI ABX INP cells were isolated as described in Supplementary Figure S4. Log2 fold change of OR expression in NFI ABX vs control INP cells shows a significant decrease in expression of zone 4-5 ORs (right).

Finally, we explored the effects of triple NFI deletion on the polygenic transcription of ORs in INP cells. We used a FACS-based strategy to isolate INPs from the ventral MOE followed by bulk RNA-seq as described earlier (Supplementary Figure S4A). Again, as with the results from ChIP-seq and Hi-C experiments, we detect a conversion toward the signatures observed in dorsomedial INPs, i.e., detection of only dorsal and dorsomedial ORs and depletion of ventral OR identities from the INP transcriptome (Figure 6C). Thus, our data reveal an unexpected correlation between OR transcription in INP cells, and two diametrically opposing gene expression outcomes in OSNs: silencing for the majority of the co-transcribed OR alleles and singular choice for one of them. We devised a genetic strategy that would test the hypothesis that polygenic OR transcription is a pre-requisite for singular OR choice.

Early OR transcription promotes OR gene choice in mOSNs

We manipulated OR transcription using a genetically modified Olfr17 allele with a tetO promoter inserted immediately downstream of its transcription start site[48]. This allele enables strong transcriptional activation of Olfr17 from the endogenous genomic locus under the control of tTA (Figure 7A, Supplementary Figure S7A). Olfr17 expression is monitored by an IRES-GFP reporter inserted immediately downstream of the Olfr17 translational stop codon (Figure 7A). To induce transcription of this “tetO-Olfr17” OR allele in INPs and iOSNs, we used Gng8-tTA transgenic mice. Gng8 is expressed in INPs and iOSNs, and completely shuts off in mOSNs (Supplementary Figure S7D). Consistent with the expression properties of Gng8 and previous reports[49, 50], we only detect GFP in the basal MOE layers of Gng8-tTA; tetO-GFP mice (Figure 7B), which are enriched for INP and iOSN cells. However, when we cross the same Gng8-tTA driver to tetO-Olfr17 mice, we detect widespread GFP signal in apical MOE layers, which contain predominantly mOSNs (Figure 7B). Since there is no tTA expression in mOSNs, we reasoned that the INP/iOSN-induced tetO-Olfr17 allele is chosen for expression by the endogenous transcriptional machinery responsible for singular OR choice. Indeed, Hi-C experiments of these OSNs revealed that Greek Islands, the intergenic OR enhancers that converge over the chosen OR allele[12-14], are recruited specifically to the tetO-Olfr17 allele (Figure 7C), explaining the sustained expression of this OR in mOSNs. The hallmark of OR choice is the singular and stable expression of the chosen allele. Consistent with this, cells expressing the tetO-Olfr17 allele do not express any other OR genes (Supplementary Figure S7E). Furthermore, treating tetO-Olfr17; Gng8-tTA mice with high doxycycline (200 mg/kg in food) for 35 days fails to extinguish tetO-Olfr17 expression in mOSNs (Supplementary Figure S7B,C). Together these findings support the notion that transcriptional induction of Olfr17 in INPs/iOSNs signals for the preferential choice of this OR in mOSNs.

Genetic induction of OR transcription in olfactory progenitors determines OR choice in mOSNs

(A) Genetic strategy for transcriptional induction of OR Olfr17 (a zone 2 identity OR) from its endogenous genomic locus. A genetically modified “tetO-Olfr17” allele contains a tetO promoter immediately downstream of the endogenous Olfr17 promoter and an IRES GFP reporter after the coding sequence[48]. In the presence of tTA a high level of tetO-Olfr17 is induced from the tetO promoter (top), while in the presence of a high amount of doxycycline (DOX) tTA is inhibited and transcription is regulated by the endogenous promoter. See also Supplementary Figure S7A for information on the genomic locus of this Olfr17 allele.

(B) tTA driven by the Gng8 promoter is expressed in INP and iOSN cells in the MOE[57]. When Gng8-tTA drives expression of a tetO-GFP allele, transcription is detected only in progenitor cells located on the basal side of the MOE, where the tTA is expressed (left)[50]. In contrast, when Gng8-tTA drives expression of tetO-Olfr17, expression persists in mature OSNs where tTA is no longer present (right). See also Supplementary Figure S7B,C for the sustained and widespread expression of the tetO-Olfr17 allele after 35 days of high DOX treatment and Supplementary Figure S7D for Gng8 expression during OSN differentiation.

(C) In situ Hi-C in tetO-Olfr17 expressing cells shows enriched contacts with interchromosomal olfactory receptor (“Greek Island”) enhancers over the Olfr17 locus, suggesting tetO-Olfr17+ OSNs are using endogenous mechanisms to sustain Olfr17 expression after Gng8-tTA is no longer present.

(D) tetO-Olfr17 expression in coronal sections of the MOE determined by GFP fluorescence. In the absence of tTA tetO-Olfr17 expression occurs only in zone 2 of the MOE (right); with high tTA induction in progenitor cells tetO-Olfr17 expression occurs throughout all zones of the MOE (left); and with low tTA induction in progenitor cells, due to the addition of a low amount of doxycycline, tetO-Olfr17 expression occurs in zone 2 and spreads dorsally to zone 1 (middle) only. Magnified views show tetO-Olfr17 expression in its native zone 2 (i) and ectopic expression in the most ventral zone 5 (ii). Mice on low DOX treatment were provided doxycycline at 1ug/ml in water throughout gestation and postnatal life.

(E) Quantification of tetO-Olfr17 expression (determined by GFP fluorescence in immunofluorescence images) relative to a normalized zonal position (illustrated on the left) in coronal sections of the MOE from tetO-Olfr17 without tTA driver (bottom), tetO-Olfr17 with Gng8-tTA driver (top), and tetO-Olfr17 with Gng8-tTA driver on low DOX (middle). 6 sections from two replicates were analyzed for tetO-Olfr17 with Gng8-tTA; 9 sections from two replicates were analyzed from tetO-Olfr17 with Gng8-tTA and low DOX; 29 sections from two replicates were analyzed for tetO-Olfr17 without tTA. The plot displays a maximum of 1000 cells randomly selected for each condition.

(F) H3K9me3 native ChIP signal over the Olfr17 locus in mOSNs from dorsal (red), dorsomedial (green) and ventral (blue) MOE shows a higher level of heterochromatin in ventral MOE.

(G) Model of OR choice in each zone of the MOE, regulated by the interplay of low-level polygenic OR transcription in INP cells, which defines the OR repertoire that can be chosen in each zone, and heterochromatic silencing, which prevents ectopic expression of more dorsal ORs. Both polygenic OR transcription in INP cells and heterochromatin deposition are influenced by NFI A, B and X transcription factors, expressed in a dorsal-low ventral-high gradient across the MOE.

Intriguingly, transient induction of Olfr17 transcription promotes preferential choice of this OR throughout the MOE, rather than only in zone 2, where Olfr17 is normally expressed (Figure 7D). In fact, the vast majority of mOSNs from zones 1 to 4 are GFP+, and only in the ventral-most zone 5 we detect a more sporadic pattern of ectopic Olfr17 choice (Figure 7D, E). We hypothesized that reduced frequency of ectopic Olfr17 expression in the most ventral segment reflects the fact that heterochromatin levels and genomic compartmentalization of this dorsomedial OR allele is highest at this MOE segment (Figure 7F). This immediately suggests that the balance between transcriptional activation and heterochromatic silencing during INP to iOSN transition determines whether an OR can be chosen for singular expression. If this hypothesis is correct, then reducing Olfr17 transcription in INP/iOSN cells should preferentially prohibit ectopic Olfr17 expression in ventral MOE segments, where heterochromatic silencing is stronger. To test this, we pharmacologically manipulated tTA activity using a low level of doxycycline (1ug/ml in water) administered throughout gestation and postnatal life of the mouse (Figure 7D), which reduces but does not eliminate tTA-driven transcription. Remarkably, mice that were subjected to this doxycycline regimen continue to frequently express Olfr17 in dorsal mOSNs (zones 1-2), but not in mOSNs from more ventral MOE segments (zones 3-5) (Figure 7D, E), where heterochromatin levels on this OR allele are highest (Figure 7F). Thus, we can manipulate the zonal expression of an OR allele in mOSNs, by pharmacologically modulating the frequency and levels of transcriptional activation in INP/iOSN cells.

Discussion

We uncovered a mechanism by which a probabilistic transcriptional process becomes skewed towards specific outcomes, transforming the relative position of a neuron across the dorsoventral axis of the MOE into biased OR gene choice. The solution to the perplexing segmentation of the MOE into distinct and reproducible territories of OR expression may be the following: polygenic OR transcription in neuronal progenitors highlights a small group of ORs that can be chosen for singular expression later in development (Figure 7G). In each MOE segment this OR mixture includes ORs that should be expressed in mOSNs of the segment, as well as ORs that are only expressed in more dorsal MOE segments (Figure 7G). As these progenitor cells differentiate into iOSNs, heterochromatic silencing may preferentially decommission from this mixture more dorsal ORs, and with lower efficiency, ORs that could be expressed in the segment, biasing this singular choice towards a spatially appropriate OR repertoire (Figure 7G). Our scRNA-seq analysis revealed two vectors in the determination of the OR ensemble that is co-transcribed in each OSN progenitor: chance, as every OR combination is unique, and determinism, as the overall zonal identities of the co-transcribed OR mixtures are informed by the position of the progenitor cell. Similarly, analysis of OR gene compartments with Hi-C revealed that genomic silencing also follows skewed patterns, eliminating preferentially ORs with more dorsal expression signatures than ORs that could be expressed in each zone, while Dip-C suggests that OR compartmentalization also retains an element of skewed randomness. The final product of these opposing probabilistic “rheostats” may be the generation of gene expression programs that do not have sufficient resolution to determine which OR will be chosen in every OSN but are precise enough to generate reproducible expression territories for each one of the ∼1400 OR genes. Recent reports describing opposing effects of chromatin compaction and transcriptional activation in the probabilistic expression patterns of the Spineless gene in R7 photoreceptors argue for the generality of this regulatory principle in vertebrates and invertebrates[51].

We identified gradients of transcription factors NFI A, B, and X as partial orchestrators of the dorsoventral patterning of OR expression, which they establish as follows: they contribute to the silencing of some dorsomedial (primarily zone 3) ORs; they activate both polygenic transcription and silencing of ventral (zone 4 and 5) ORs; and they have no influence on the expression of dorsal-most (class I and zone 1) ORs. Without NFI transcription factors, the majority of the MOE, excluding dorsal-most zone1, defaults to a zone 2 identity. Given that NFI factors are predominantly known as regulators of embryonic and adult stem cell biology [44, 52], it is surprising that in the olfactory system their deletion does not interfere with the maintenance of stem cell populations, but with the OR expression patterns in post-mitotic, fully differentiated mOSNs. Interestingly, triple NFI deletion after the onset of singular OR choice has no effect in OR patterning, consistent with the emerging model that OR specification takes place exclusively at the INP to iOSN transition, and the notion that these patterning factors are not required for maintenance of OR transcription. Thus, we speculate that singular OR gene choice in OSNs can be executed by the common nucleoprotein complex of Lhx2/Ebf/Ldb1 bound to the multi-enhancer hub, consistent with the fact that we detect hubs of similar constitution associating with active ORs in different zones [14].

A question emerging from these observations is why not use the same transcription factor gradients to regulate both polygenic and monogenic OR transcription? The answer is likely related to the absolute requirement for transcriptional singularity: transcription factor gradients can transcribe specific OR mixtures in a DV-responsive fashion, but they cannot activate only a single OR promoter among the many they can bind to. But even if singularity was achievable by transcription factor combinations and the OR-elicited feedback, OR promoters with the strongest binding motifs would be consistently chosen first, excluding ORs with weaker promoters in a “winner takes all” model. This would result in preferential choice of specific ORs, reduced diversity in OR representation, and a narrower sensory spectrum for the olfactory system. With the process revealed here, any OR promoter activated in INPs/iOSNs is probabilistically chosen for singular expression in OSNs. Thus, by segregating OR gene regulation into two stages, polygenic transcription in progenitor cells and singular choice in OSNs, the olfactory system can impose deterministic biases while assuring equitable receptor representation. Of course, this system has limitations in preserving transcriptional equity: artificial transcriptional induction of an OR allele in OSN progenitors under the powerful tetO promoter bypasses these constraints and results in biased choice of this allele in most mOSNs. This immediately suggests that cis OR regulatory elements are subject to selective pressure that preserves their weak transcriptional activation properties, explaining why robust OR transcription in mOSNs requires assembly of interchromosomal multi-enhancer hubs.

In this note, zones may also have evolved to satisfy the requirement for distributed OR representation: if dorsal-most ORs, which are detected in every OSN progenitor regardless of DV origin, have the most frequently activated promoters, then silencing them in more ventral MOE segments assures that other OR identities will also have the chance to be expressed. Consistent with this model is the observation that mutations on the Lhx2 or Ebf binding sites of the promoter of dorsal OR M71 result in less frequent and more ventral M71 expression patterns [53]. Thus, DV segmentation of the MOE may serve as a mechanism that prevents ORs with stronger differences in promoter strength from competing for singular expression, assuring that every OR is expressed at meaningful, for odor perception, frequencies. In addition, as our spatial transcriptomic data showed, zonal regulation assures that ORs are expressed in a reproducibly patterned fashion in the MOE. While in wild type mice unbiased machine learning approaches identify at least 5 distinct OR expression patterns, in the triple NFI cKO mice these patterns become intermixed for all but zone 1 ORs. With recent observations arguing that individual mitral cells, the second order neurons in the olfactory circuit, have patterned projections in the brain [54], non-random OR expression in the MOE may contribute to putative hardwired components of odor perception and valence [55].

Polygenic OR transcription as the arbiter between OR gene silencing and OR gene choice

A peculiar feature of the OR gene family that had emerged from our past work is that OR gene silencing is highest in the very cells that express ORs [20]. Our zonal analysis further strengthened this intriguing correlation, as both H3K9me3/H3K79me3 and genomic compartmentalization in each MOE segment are strongest on OR groups that are transcriptionally active during OSN differentiation. A fascinating implication of this observation is that early OR transcription is the signal for both genomic silencing and singular choice. Although the former is only implied by the strong correlation between OR transcription in OSN progenitors and genomic silencing, the latter is experimentally supported by the striking observation that strong transcriptional induction of Olfr17 at the INP/iOSN stage results in strong recruitment of the Greek Island hub, and stable choice of this OR allele in most mOSNs throughout the MOE. Such a mechanism of promoter choice influenced by spatially-determined early transcription could also explain the recent demonstration that clustered Pcdh choice, which is regulated by anti-sense transcription [56], abides by spatial restrictions in the neocortex [6].

How could two fundamentally opposite gene expression outcomes be encoded on the same molecular feature? We propose that the timing and levels of transcriptional induction could be the arbiters between genomic silencing and singular choice. ORs that are transcribed first in the INP stage, when the Greek Island hub cannot yet form due to the continuous expression of Lamin b receptor [21], are most likely to be silenced. OR alleles activated during the assembly of the multi-enhancer hub, at the INP to iOSN transition, may compete for hub recruitment. The OR allele that will first associate with a multi-enhancer hub will be stably protected from heterochromatic silencing, possibly due to the significantly increased rates of OR transcription, whereas the other co-transcribed ORs will succumb to heterochromatic silencing. If timing and rates of OR transcription determine whether an OR allele will be silenced or chosen, then an OR allele that is highly transcribed in both INP and iOSN stage should evade silencing and dominate the competition for hub recruitment, explaining the striking expression pattern of the tTA-induced Olfr17 allele. Thus, according to this model, in each OSN ORs with more dorsal identity will be transcribed first, because they have stronger promoters, and therefore will become silenced in higher frequency; ORs with the correct zonal identity will be transcribed later, with a chance to associate with the Greek Island hub, explaining why one is chosen and the rest are silenced; ORs with more ventral identities will not be transcribed at all, thus, will not be silenced but also will not be chosen. In other words, singular OR transcription may not depend on the silencing of every single OR in the genome: by encoding silencing and stable choice with the same exact molecular feature, OSNs choose one and silence a small fraction of the whole OR repertoire in each nucleus—the rest are not relevant. Notably, this constitutes a refinement of our original model, which proposed that all but one OR alleles become silenced in each OSN[20]. Without knowledge of the zonal identity of most ORs, and without the technical ability to perform Hi-C and ChIP-seq on micro-dissected MOE zones, the striking patterns of gradual increase of OR silencing along the D-V axis of the MOE were not appreciated. As the cellular and temporal resolution of our genome wide approaches increases, the aforementioned model will be further refined.

Limitations of this study

Our experiments did not clarify whether NFI proteins bind directly on OR promoters, or act indirectly by activating other transcription factors and chromatin modifying enzymes. Although there is a statistically significant enrichment of NFI motifs on zone 4/5 OR promoters compared to the other OR promoters (data not shown), we were not able to detect direct binding of NFI proteins on these promoters, which is expected since these promoters are active in less than 1% of the cells. Given that our studies provide the mechanism by which NFI gradients establish zonal boundaries, via polygenic OR transcription and chromatin-mediated silencing, answering this question is not essential for understanding the mechanism of dorsoventral patterning of OR expression. A second limitation of this study is that it did not reveal the mechanisms that regulate the expression of the dorsal-most ORs (zone 1 ORs), as NFI deletion had no effects on the expression and chromatin regulation of these OR genes. However, having revealed the regulatory logic whereby these patterns are established, we expect that other transcription factors with zonal expression patterns identified here regulate early transcription and silencing of these genes across the MOE.

Supporting information